---
title: "Genotype QC — pre-imputation report"
date: today
date-format: "YYYY-MM-DD"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-expand: 2
    number-sections: true
    embed-resources: true
    fig-align: center
    tbl-cap-location: bottom
    theme: cosmo
    grid:
      body-width: 1000px
      sidebar-width: 260px
      margin-width: 140px
execute:
  echo: false
  warning: false
  message: false
params:
  run_dir: ""
  prefix: ""
  config_json: ""
  cohort: ""
---

<!--
=============================================================================
genotype_qc_report.qmd


Renders a standalone HTML report from a genotype_qc_preimputation run, 
reading the Cromwell run directory.

USAGE
  quarto render src/genotype_qc_report.qmd \
      -P run_dir=/path/to/cromwell-executions/genotype_qc_preimputation/<uuid> \
      -P cohort="My cohort" \
      -P config_json=config/genotype_qc_preimputation_inputs.json \
      --output-dir /absolute/path/for/the/html

PARAMETERS
  run_dir      the workflow UUID directory holding the call-* subdirectories
  cohort       free-text cohort name shown in the header (optional)
  config_json  workflow inputs JSON, to document the thresholds (optional)
  prefix       output prefix; auto-detected from *_pipeline.log if omitted

RUN-DIRECTORY SPECIFICS
  - `call-<Task>/attempt-N/execution` appears when a task is retried
    (`maxRetries: 1` is set on every task); the highest attempt wins.
  - `execution/glob-*/` holds duplicates of globbed outputs (the VCFs) and
    `call-<Task>/inputs/` holds copies of upstream outputs. Both would
    double-count, so listings are never recursive.
  - The run directory also holds every intermediate dataset, each task's plink
    log, stdout/stderr and exit code. Those are summarised in the inventory.
=============================================================================
-->

```{r setup}
#| include: false

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(stringr)
  library(knitr)
})

knitr::opts_chunk$set(
  dev       = "png",
  dpi       = 150,
  fig.width = 8,
  fig.retina = 2
)

## ---------------------------------------------------------------------------
## Palette
##
## Two categorical slots only (blue / orange). Validated on a white surface for
## all pairs: CVD deltaE 24.7, normal-vision deltaE 33.6, both above 3:1
## contrast. Charts that would need more than two hues use small multiples or a
## single hue with the categories on an axis instead, so identity is never
## carried by colour alone. Grey is reserved for context/"other", never a series.
## ---------------------------------------------------------------------------
PAL <- c(blue = "#2a78d6", orange = "#eb6834", grey = "#8a8a86",
         ink = "#0b0b0b", ink2 = "#52514e", grid = "#e6e6e3")

theme_report <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 12, colour = PAL[["ink"]]),
      plot.subtitle    = element_text(size = 10, colour = PAL[["ink2"]]),
      plot.caption     = element_text(size = 8,  colour = PAL[["ink2"]], hjust = 0),
      axis.title       = element_text(size = 10, colour = PAL[["ink2"]]),
      axis.text        = element_text(size = 9,  colour = PAL[["ink2"]]),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = PAL[["grid"]], linewidth = 0.3),
      strip.text       = element_text(face = "bold", size = 10, colour = PAL[["ink"]]),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = 9, colour = PAL[["ink2"]])
    )
}
theme_set(theme_report())

## ---------------------------------------------------------------------------
## Locating files inside the Cromwell run directory
## ---------------------------------------------------------------------------
## Quarto renders with the working directory set to the .qmd's own directory,
## so a relative path given on the command line is interpreted from src/.
## Try it as given, then walk up towards the project root. A "~" reaches R
## unexpanded when the shell did not expand it, so do that here too.
resolve_path <- function(p) {
  if (!nzchar(p)) return(p)
  p <- path.expand(p)
  for (cand in c(p, file.path("..", p), file.path("..", "..", p))) {
    if (file.exists(cand) || dir.exists(cand)) return(normalizePath(cand, mustWork = FALSE))
  }
  normalizePath(p, mustWork = FALSE)
}

RD <- resolve_path(params$run_dir)
if (!nzchar(params$run_dir) || !dir.exists(RD)) {
  stop("run_dir not found: ", RD, "\n",
       "  Point it at the workflow UUID directory (the one holding call-* dirs)\n",
       "  and give an ABSOLUTE path — Quarto renders with the working directory\n",
       "  set to the .qmd's own directory, not the one you invoked it from:\n\n",
       "    quarto render /path/to/genotype_qc_pipeline/src/genotype_qc_report_cromwell.qmd \\\n",
       "        -P run_dir=/absolute/path/to/cromwell-executions/genotype_qc_preimputation/<uuid> \\\n",
       "        --output-dir /absolute/path/for/the/html\n")
}

call_dirs <- sort(list.dirs(RD, recursive = FALSE, full.names = FALSE))
call_dirs <- call_dirs[str_starts(call_dirs, "call-")]
if (length(call_dirs) == 0) {
  stop("no call-* directories under ", RD,
       "\n  This does not look like a Cromwell run directory.")
}
tasks <- str_remove(call_dirs, "^call-")

## A retried task gets call-<Task>/attempt-N/execution alongside the original
## call-<Task>/execution. The last attempt is the one that produced the outputs.
exec_dir <- function(task) {
  base <- file.path(RD, paste0("call-", task))
  if (!dir.exists(base)) return(NA_character_)
  att <- list.dirs(base, recursive = FALSE, full.names = TRUE)
  att <- att[str_detect(basename(att), "^attempt-\\d+$")]
  if (length(att)) {
    n <- as.integer(str_extract(basename(att), "\\d+"))
    d <- file.path(att[which.max(n)], "execution")
    if (dir.exists(d)) return(d)
  }
  d <- file.path(base, "execution")
  if (dir.exists(d)) d else NA_character_
}

## Which task wrote each thing the report asks for. The (subdir, pattern) keys
## are the ones used by the collected-layout report, so the body of the two
## documents stays identical; `rt` is the name the file actually has in the run
## directory when it differs from its collected name.
R_ <- function(tasks, rt = NULL) list(tasks = tasks, rt = rt)
ROUTES <- list(
  "|_pipeline\\.log$"                            = R_("WriteLog"),

  "plots|threshold_sweep\\.png$"                 = R_("ThresholdSweep"),
  "plots|mac_plot\\.png$"                        = R_("MacFilter"),
  "plots|heterozygosity\\.png$"                  = R_("HeterozygosityCheck"),
  "plots|_ancestry_pca\\.png$"                   = R_("AncestryPCA"),

  "reports|initial_variants_per_chr\\.txt$"      = R_("InitialVariantsPerChromosome",
                                                      "^variants_per_chr_report\\.txt$"),
  "reports|final_variants_per_chr\\.txt$"        = R_("PostImputationVariantsPerChromosome",
                                                      "^variants_per_chr_report\\.txt$"),
  "reports|_annotated\\.sexcheck$"               = R_("SexCheck"),
  "reports|\\.sexcheck$"                         = R_("SexCheck"),
  "reports|sex_fail_samples\\.fam$"              = R_("RemoveSexFails", "\\.fam$"),
  "reports|^freq_before\\.frq$"                  = R_("MacFilter"),
  "reports|^freq_after\\.frq$"                   = R_("MacFilter"),
  "reports|\\.hwe$"                              = R_("HweFilterSubset"),
  "reports|R_check\\.het$"                       = R_("HeterozygosityCheck"),
  "reports|het_fail_samples\\.txt$"              = R_("HeterozygosityCheck",
                                                      "^het_fail_ind\\.txt$"),
  "reports|relatedness_flagged_samples\\.txt$"   = R_("RelatednessCheck",
                                                      "king\\.cutoff\\.out\\.id$"),
  "reports|relatedness_unrelated_samples\\.txt$" = R_("RelatednessCheck",
                                                      "king\\.cutoff\\.in\\.id$"),
  "reports|\\.genome$"                           = R_("RelatednessCheck"),
  "reports|check-bim\\.log$"                     = R_("PrepareForImputation"),
  "reports|_sample_qc_status\\.tsv$"             = R_("CreateQCStatusTable"),
  "reports|ancestry_assignments\\.tsv$"          = R_("AncestryPCA"),

  "pca|ancestry_assignments\\.tsv$"              = R_("AncestryPCA"),
  "pca|\\.eigenval$"                             = R_(c("AncestryPCA", "CovariatePCA")),
  "pca|_covariate_pca_covariates\\.tsv$"         = R_("CovariatePCA", "_covariates\\.tsv$"),
  "pca|_covariates\\.tsv$"                       = R_("CovariatePCA"),
  "pca|_covariate_pca\\.eigenval$"               = R_("CovariatePCA", "\\.eigenval$"),

  ## PrepareForImputation also holds the per-chromosome *-updated-chr* filesets,
  ## so the combined dataset has to be matched explicitly.
  "qc_dataset|\\.bed$"                           = R_("PrepareForImputation", "_combined\\.bed$"),
  "qc_dataset|\\.bim$"                           = R_("PrepareForImputation", "_combined\\.bim$"),
  "qc_dataset|\\.fam$"                           = R_("PrepareForImputation", "_combined\\.fam$"),
  "qc_dataset|\\.pgen$"                          = R_("CombinedToPlink2"),
  "qc_dataset|\\.pvar$"                          = R_("CombinedToPlink2"),
  "qc_dataset|\\.psam$"                          = R_("CombinedToPlink2"),

  "vcfs|\\.vcf\\.gz$"                            = R_("PrepareForImputation",
                                                      "_chr[0-9]+\\.vcf\\.gz$"),
  "vcfs|\\.vcf\\.gz\\.tbi$"                      = R_("PrepareForImputation",
                                                      "_chr[0-9]+\\.vcf\\.gz\\.tbi$"),
  "vcfs|\\.tbi$"                                 = R_("PrepareForImputation",
                                                      "_chr[0-9]+\\.vcf\\.gz\\.tbi$"),

  "subset|_keep\\.txt$"                          = R_("SubsetUnrelated"),
  "subset|\\.bed$"                               = R_("SubsetUnrelated"),
  "subset|\\.bim$"                               = R_("SubsetUnrelated"),
  "subset|\\.fam$"                               = R_("SubsetUnrelated"),
  "subset|\\.pgen$"                              = R_("SubsetToPlink2"),
  "subset|\\.pvar$"                              = R_("SubsetToPlink2"),
  "subset|\\.psam$"                              = R_("SubsetToPlink2"),
  "subset|_covariates\\.tsv$"                    = R_("SubsetCovariatePCA"),
  "subset|\\.eigenval$"                          = R_("SubsetCovariatePCA")
)

## Fallback for any (subdir, pattern) not in the table: search the tasks that
## feed that part of the collected layout.
SUBDIR_TASKS <- list(
  "plots"      = c("ThresholdSweep", "MacFilter", "HeterozygosityCheck", "AncestryPCA"),
  "reports"    = c("SexCheck", "RemoveSexFails", "InitialVariantsPerChromosome",
                   "MacFilter", "HweFilterSubset", "HeterozygosityCheck",
                   "RelatednessCheck", "AncestryPCA", "CreateQCStatusTable",
                   "PrepareForImputation", "PostImputationVariantsPerChromosome"),
  "pca"        = c("AncestryPCA", "CovariatePCA"),
  "qc_dataset" = c("PrepareForImputation", "CombinedToPlink2"),
  "vcfs"       = "PrepareForImputation",
  "subset"     = c("SubsetUnrelated", "SubsetToPlink2", "SubsetCovariatePCA")
)

`%||%` <- function(a, b) if (is.null(a)) b else a

## Never recursive: glob-*/ duplicates the VCFs and inputs/ duplicates upstream
## outputs, and both would be counted twice.
picks <- function(subdir, pattern) {
  r <- ROUTES[[paste0(subdir, "|", pattern)]]
  if (is.null(r)) {
    fallback <- if (!nzchar(subdir)) "WriteLog" else SUBDIR_TASKS[[subdir]] %||% tasks
    r <- R_(fallback)
  }
  pat <- if (is.null(r$rt)) pattern else r$rt
  out <- character(0)
  for (tk in r$tasks) {
    d <- exec_dir(tk)
    if (is.na(d)) next
    out <- c(out, list.files(d, pattern = pat, full.names = TRUE))
  }
  sort(unique(out))
}

pick <- function(subdir, pattern) {
  h <- picks(subdir, pattern)
  if (length(h) == 0) NA_character_ else h[1]
}
have <- function(p) length(p) == 1 && !is.na(p) && file.exists(p)

rel  <- function(p) sub(paste0("^", RD, .Platform$file.sep), "", p, fixed = FALSE)
fref <- function(p) if (have(p)) paste0("`", rel(p), "`") else "_(not produced)_"

nlines <- function(p) if (have(p)) length(read_lines(p, progress = FALSE)) else NA_integer_
fsize  <- function(p) if (have(p)) file.info(p)$size else NA_real_

## Non-breaking space so "512 KB" never wraps across two table lines
hsize <- function(b) {
  ifelse(is.na(b), "—",
    ifelse(b >= 1024^3, sprintf("%.1f GB", b / 1024^3),
      ifelse(b >= 1024^2, sprintf("%.1f MB", b / 1024^2),
        ifelse(b >= 1024,   sprintf("%.0f KB", b / 1024), sprintf("%.0f B", b)))))
}

## ---------------------------------------------------------------------------
## Output helpers (used inside results="asis" chunks)
## ---------------------------------------------------------------------------
n_fmt <- function(x) ifelse(is.na(x), "—", formatC(as.numeric(x), format = "d", big.mark = ","))
p_fmt <- function(x) ifelse(is.na(x), "—", sprintf("%.1f%%", 100 * x))

show_table <- function(df, caption = NULL, align = NULL) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  ## Sample and family identifiers are numeric in PLINK files but are not
  ## quantities — keep them out of the thousands-separator formatting.
  id_cols <- c("FID", "IID", "FID1", "IID1", "FID2", "IID2", "ID", "PC")
  df <- df |> mutate(across(any_of(id_cols), as.character))
  cat(knitr::kable(df, caption = caption, align = align,
                   format.args = list(big.mark = ",")), sep = "\n")
  cat("\n\n")
}

## Plots produced by the pipeline live outside the report's directory, so they
## are inlined as base64 data URIs rather than linked — this keeps the rendered
## HTML a single self-contained file that can be emailed as-is.
show_image <- function(p, caption = "") {
  if (!have(p)) return(invisible(NULL))
  uri <- tryCatch(knitr::image_uri(p), error = function(e) p)
  cat(sprintf("![%s](%s)\n\n", caption, uri))
}

absent <- function(msg) {
  cat("\n::: {.callout-warning appearance='simple'}\n**Not available in this run.** ",
      msg, "\n:::\n\n", sep = "")
}

note <- function(msg) {
  cat("\n::: {.callout-note appearance='simple'}\n", msg, "\n:::\n\n", sep = "")
}

collapsible <- function(title, text) {
  cat("\n::: {.callout-note collapse='true'}\n## ", title, "\n\n```\n", sep = "")
  cat(text, sep = "\n")
  cat("\n```\n:::\n\n")
}

## ---------------------------------------------------------------------------
## Task exit codes. A run directory can hold a failed or half-finished run, so
## this is checked before anything is reported as a result.
## ---------------------------------------------------------------------------
task_status <- tibble(task = tasks) |>
  mutate(
    dir = vapply(task, exec_dir, ""),
    rc  = vapply(dir, function(d) {
      f <- file.path(d, "rc")
      if (!is.na(d) && file.exists(f)) str_trim(read_lines(f, progress = FALSE)[1]) else NA_character_
    }, "")
  )
failed_tasks <- task_status |> filter(!is.na(rc) & rc != "0")
unfinished   <- task_status |> filter(is.na(rc))

## ---------------------------------------------------------------------------
## The pipeline log — the backbone of the report
## ---------------------------------------------------------------------------
log_file <- pick("", "_pipeline\\.log$")
log_txt  <- if (have(log_file)) read_lines(log_file, progress = FALSE) else character(0)

prefix <- params$prefix
if (!nzchar(prefix) && have(log_file)) {
  prefix <- sub("_pipeline\\.log$", "", basename(log_file))
}

run_version <- str_match(paste(log_txt, collapse = "\n"),
                         "genotype_qc_preimputation\\s+v(\\S+)")[, 2]
run_date    <- str_match(paste(log_txt, collapse = "\n"),
                         "Run date\\s*:\\s*(\\S+ \\S+)")[, 2]

## Step lines carrying "SNPs: n  Samples: n"
step_counts <- tibble(line = log_txt) |>
  filter(str_detect(line, "SNPs:\\s*\\d+\\s+Samples:\\s*\\d+")) |>
  mutate(
    step    = str_trim(str_extract(line, "^.*?(?=\\s+SNPs:)")),
    snps    = as.integer(str_match(line, "SNPs:\\s*(\\d+)")[, 2]),
    samples = as.integer(str_match(line, "Samples:\\s*(\\d+)")[, 2])
  ) |>
  select(step, snps, samples) |>
  mutate(order = row_number())

first_step <- if (nrow(step_counts)) slice(step_counts, 1) else NULL
## last step: last line in the output log, might not be what want
# last_step  <- if (nrow(step_counts)) slice(step_counts, n()) else NULL

## define last step by step number, rather than the last line in the output log
last_step <- step_counts |>
  filter(str_detect(step, regex("^Step 1[01]\\b"))) |>
  slice_tail(n = 1)
if (nrow(last_step) == 0) last_step <- if (nrow(step_counts)) slice(step_counts, n()) else NULL


## Free-text log lines that are not SNP/sample counts (ancestry, LD, PCA, ...)
log_notes <- log_txt[str_detect(log_txt, "^Step ") &
                     !str_detect(log_txt, "SNPs:\\s*\\d+\\s+Samples:")]

## ---------------------------------------------------------------------------
## Common tables read once and reused
## ---------------------------------------------------------------------------
qc_file  <- pick("reports", "_sample_qc_status\\.tsv$")
qc_tbl   <- if (have(qc_file)) suppressWarnings(read_tsv(qc_file, show_col_types = FALSE,
                                                         progress = FALSE)) else NULL

anc_file <- pick("reports", "ancestry_assignments\\.tsv$")
if (!have(anc_file)) anc_file <- pick("pca", "ancestry_assignments\\.tsv$")
anc_tbl  <- if (have(anc_file)) suppressWarnings(read_tsv(anc_file, show_col_types = FALSE,
                                                          progress = FALSE)) else NULL

pc_cols <- function(df) if (is.null(df)) character(0) else grep("^PC\\d+$", names(df), value = TRUE)

## Sample lists come in three shapes: PLINK 2 `.id` files with a "#FID IID"
## header, the pipeline's own lists with a plain "FID IID" header, and bare
## two-column lists with no header at all.
read_id_file <- function(p) {
  if (!have(p)) return(NULL)
  x <- read_lines(p, progress = FALSE)
  x <- x[nzchar(str_trim(x))]
  if (length(x) == 0) return(tibble(FID = character(), IID = character()))
  if (str_detect(x[1], "^#") || str_detect(x[1], "^\\s*#?FID\\b")) x <- x[-1]
  if (length(x) == 0) return(tibble(FID = character(), IID = character()))
  parts <- str_split(str_trim(x), "\\s+", simplify = TRUE)
  if (ncol(parts) == 1) tibble(FID = parts[, 1], IID = parts[, 1])
  else tibble(FID = parts[, 1], IID = parts[, 2])
}

## Variants-per-chromosome report -> tidy table.
## Lines look like "  Chromosome  1:   48604 variants" or
## "  Chromosome 23 (X):   26957 variants" — keep the parenthetical annotation.
read_varchr <- function(p) {
  if (!have(p)) return(NULL)
  x <- read_lines(p, progress = FALSE)
  m <- str_match(x, "Chromosome\\s+(\\d+)\\s*(\\([A-Za-z]+\\))?\\s*:\\s+(\\d+)\\s+variants")
  ok <- !is.na(m[, 1])
  if (!any(ok)) return(NULL)
  ## Short axis labels: "23 (X)" collides with its neighbours on a 26-bar axis.
  tibble(chr = as.integer(m[ok, 2]), tag = m[ok, 3],
         variants = as.integer(m[ok, 4])) |>
    arrange(chr) |>
    mutate(chr_label = ifelse(is.na(tag), as.character(chr),
                              str_remove_all(tag, "[()]"))) |>
    select(chr, chr_label, variants)
}
```

```{r headline}
#| results: asis

## Assemble only the fields this run actually has — with no pipeline log there
## is no prefix, version or run date to show.
bits <- character(0)
if (nzchar(params$cohort))  bits <- c(bits, paste0("**Cohort:** ", params$cohort))
if (nzchar(prefix))         bits <- c(bits, paste0("**Run:** `", prefix, "`"))
if (!is.na(run_version))    bits <- c(bits, paste0("**pipeline version** ", run_version))
if (!is.na(run_date))       bits <- c(bits, paste0("**run date** ", run_date))
bits <- c(bits, paste0("**Cromwell run** `", basename(RD), "`"))

cat("::: {.callout-tip appearance='simple'}\n")
cat(paste(bits, collapse = "  ·  "), "\n", sep = "")
cat(":::\n\n")

## Anything other than "every task exited 0" is stated before the results, not
## after them — a run directory can hold a failed or still-running workflow.
if (nrow(failed_tasks)) {
  cat("\n::: {.callout-important}\n## This run did not complete cleanly\n\n",
      nrow(failed_tasks), " task(s) exited non-zero: ",
      paste(sprintf("`%s` (rc %s)", failed_tasks$task, failed_tasks$rc), collapse = ", "),
      ".\n\nEverything below is read from whatever those tasks did write, so treat ",
      "it as partial.\n:::\n\n", sep = "")
} else if (nrow(unfinished)) {
  cat("\n::: {.callout-important}\n## This run may still be in progress\n\n",
      nrow(unfinished), " task(s) have no exit-code file: ",
      paste(sprintf("`%s`", unfinished$task), collapse = ", "),
      ".\n:::\n\n", sep = "")
}
```

This report summarises one execution of the `genotype_qc_preimputation` WDL
workflow: what was filtered at each step, how many variants and samples
survived, what the diagnostic plots show, and which task in the Cromwell run
directory produced each file. Numbers are read directly from the run's own
outputs — nothing here is recomputed from the genotypes.

# Headline numbers

```{r headline-table}
#| results: asis

rows <- list()
add <- function(metric, value, source) rows[[length(rows) + 1]] <<-
  tibble(Metric = metric, Value = value, `Read from` = source)

if (!is.null(first_step)) {
  add("Variants entering QC", n_fmt(first_step$snps), rel(log_file))
  add("Samples entering QC",  n_fmt(first_step$samples), rel(log_file))
}
if (!is.null(last_step)) {
  add("Variants in the final deliverable", n_fmt(last_step$snps), rel(log_file))
  add("Samples in the final deliverable",  n_fmt(last_step$samples), rel(log_file))
}
if (!is.null(first_step) && !is.null(last_step)) {
  add("Variants retained", p_fmt(last_step$snps / first_step$snps), "derived")
  add("Samples retained",  p_fmt(last_step$samples / first_step$samples), "derived")
}

rel_out_f <- pick("reports", "relatedness_flagged_samples\\.txt$")
rel_out <- read_id_file(rel_out_f)
if (!is.null(rel_out)) add("Samples flagged as related (KING)", n_fmt(nrow(rel_out)),
                           rel(rel_out_f))

het_fail <- pick("reports", "het_fail_samples\\.txt$")
if (have(het_fail)) add("Heterozygosity outliers removed",
                        n_fmt(nrow(read_id_file(het_fail))),
                        rel(het_fail))

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  top <- anc_tbl |> count(superpop, sort = TRUE)
  add("Ancestry composition",
      paste(sprintf("%s: %s", top$superpop, n_fmt(top$n)), collapse = ", "),
      basename(anc_file))
}

vcfs <- picks("vcfs", "\\.vcf\\.gz$")
if (length(vcfs)) add("Imputation-ready VCFs", sprintf("%d files (%s)",
                       length(vcfs), hsize(sum(file.info(vcfs)$size))),
                      "call-PrepareForImputation")

add("Workflow tasks", sprintf("%d (%s)", nrow(task_status),
                              if (nrow(failed_tasks) == 0 && nrow(unfinished) == 0)
                                "all exited 0" else "see warning above"),
    paste0("call-*/execution/rc"))

if (length(rows)) {
  show_table(bind_rows(rows), align = c("l", "r", "l"))
} else {
  absent("No pipeline log was found, so headline numbers could not be derived.")
}
```

# Filtering trajectory

The workflow chains PLINK filters, writing a new `bed/bim/fam` triplet at each
step. The table below is the per-step count log the run wrote itself; the chart
shows the same numbers as a retention trajectory. Sharp drops are where QC did
its work — the interesting question at each step is whether the drop matches the
threshold that was applied.

```{r trajectory}
#| results: asis
#| fig-height: 6.5

if (nrow(step_counts) == 0) {
  absent(paste("No `*_pipeline.log` under `call-WriteLog/` — the whole trajectory",
               "section depends on it. That task runs last, so its absence usually",
               "means the workflow did not finish."))
} else {
  show_table(
    step_counts |>
      mutate(
        `Δ variants` = c(NA, diff(snps)),
        `Δ samples`  = c(NA, diff(samples))
      ) |>
      select(Step = step, Variants = snps, `Δ variants`,
             Samples = samples, `Δ samples`),
    caption = "Per-step variant and sample counts (source: the run's pipeline log)."
  )
}
```

```{r trajectory-plot}
#| fig-height: 6.5

if (nrow(step_counts) > 1) {
  traj <- step_counts |>
    pivot_longer(c(snps, samples), names_to = "metric", values_to = "n") |>
    mutate(
      metric = factor(recode(metric, snps = "Variants", samples = "Samples"),
                      levels = c("Variants", "Samples")),
      label  = str_trunc(str_squish(str_remove(step, "^Step\\s+\\S+\\s*")), 26),
      label  = factor(sprintf("%02d %s", order, label),
                      levels = sprintf("%02d %s", order, label)[!duplicated(order)])
    )

  ggplot(traj, aes(x = label, y = n, group = metric)) +
    geom_line(colour = PAL[["blue"]], linewidth = 0.7) +
    geom_point(colour = PAL[["blue"]], size = 2.2) +
    geom_text(aes(label = formatC(n, format = "d", big.mark = ",")),
              vjust = -0.9, size = 2.7, colour = PAL[["ink2"]]) +
    facet_wrap(~metric, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0.08, 0.18))) +
    labs(title = "Variants and samples retained through QC",
         subtitle = "Each panel has its own scale — the two measures are not comparable in size",
         x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
}
```

```{r trajectory-notes}
#| results: asis

if (length(log_notes)) {
  cat("Steps that report something other than a count — ancestry assignment,",
      "LD pruning, PCA, per-chromosome totals — logged the following:\n\n")
  collapsible("Additional log lines", log_notes)
}

if (length(log_txt)) collapsible("Full pipeline log", log_txt)
```

# Missingness filters (steps 1–2)

Step 1 (`--geno`) drops variants genotyped in too few samples; step 2
(`--mind`) drops samples genotyped at too few variants. Order matters: variants
are cleaned first so that a handful of failed assays cannot condemn otherwise
good samples.

The threshold sweep below is informational — the pipeline does not choose
thresholds from it. It shows how many variants and samples each candidate
threshold would have cost, so you can confirm the configured values sit on the
flat part of the curve rather than on a cliff.

```{r threshold-sweep}
#| results: asis

p <- pick("plots", "threshold_sweep\\.png$")
if (have(p)) {
  show_image(p, "Variant and sample retention across candidate --geno / --mind thresholds")
  cat("Source file: ", fref(p), "\n\n", sep = "")
  ## Only the run directory keeps the numbers behind the plot.
  gs <- pick("plots", "^geno_sweep\\.txt$")
  ms <- pick("plots", "^mind_sweep\\.txt$")
  if (have(gs) || have(ms)) {
    cat("The underlying sweep tables are ", fref(gs), " and ", fref(ms),
        " — these are not copied out by `collect_cromwell_results.bash`.\n\n", sep = "")
  }
} else {
  absent("No `threshold_sweep.png` under `call-ThresholdSweep/` — the task did not run or produced no plot.")
}
```

## Variants per chromosome, before imputation prep

```{r varchr-initial}
#| results: asis
#| fig-height: 3.6

vc_init <- read_varchr(pick("reports", "initial_variants_per_chr\\.txt$"))
if (is.null(vc_init)) {
  absent("No `variants_per_chr_report.txt` under `call-InitialVariantsPerChromosome/`.")
} else {
  cat("Counted on the post-missingness dataset. Source file: ",
      fref(pick("reports", "initial_variants_per_chr\\.txt$")), "\n\n", sep = "")
}
```

```{r varchr-initial-plot}
#| fig-height: 3.6

if (!is.null(vc_init)) {
  ggplot(vc_init, aes(x = factor(chr_label, levels = chr_label), y = variants)) +
    geom_col(fill = PAL[["blue"]], width = 0.72) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
    labs(title = "Variants per chromosome entering QC",
         x = "Chromosome", y = "Variants")
}
```

```{r varchr-initial-table}
#| results: asis

if (!is.null(vc_init)) {
  show_table(
    vc_init |> select(Chromosome = chr_label, Variants = variants) |>
      bind_rows(tibble(Chromosome = "Total", Variants = sum(vc_init$variants))),
    caption = "Variants per chromosome before imputation preparation."
  )
}
```

# Sex check (step 3)

X-chromosome heterozygosity is compared with the sex recorded in the `.fam`
file. Samples whose F-statistic contradicts the recorded sex are removed — a
mismatch usually means a sample swap, which invalidates every phenotype
attached to it. The step is skipped automatically when the array carries no X
markers.

```{r sexcheck}
#| results: asis

## Two .sexcheck files are written: PLINK's raw output and the pipeline's
## annotated copy. Prefer the annotated one.
sx <- pick("reports", "_annotated\\.sexcheck$")
if (!have(sx)) sx <- pick("reports", "\\.sexcheck$")
sx_prob <- pick("reports", "_problem_samples\\.txt$")
sx_kept <- pick("reports", "sex_fail_samples\\.fam$")

if (!have(sx)) {
  skipped <- any(str_detect(log_txt, regex("sex check.*skipped", ignore_case = TRUE)))
  if (skipped) {
    note(paste("**Sex check skipped.** The dataset contained no X-chromosome",
               "SNPs, so no F-statistic could be computed. This is expected for",
               "autosome-only arrays and does not indicate a problem."))
  } else {
    absent("No `*.sexcheck` under `call-SexCheck/`.")
  }
} else {
  ## The annotated report starts with "##" provenance lines (chromosome coding,
  ## number of X SNPs used) before the PLINK header.
  hdr <- read_lines(sx, n_max = 10, progress = FALSE)
  hdr <- str_trim(str_remove(hdr[str_starts(hdr, "##")], "^##\\s*"))
  sxt <- suppressWarnings(read_table(sx, comment = "##", show_col_types = FALSE,
                                     progress = FALSE))
  cat("Source file: ", fref(sx), "\n\n", sep = "")
  if (length(hdr)) cat(paste0("- ", hdr, collapse = "\n"), "\n\n", sep = "")

  if ("STATUS" %in% names(sxt)) {
    show_table(sxt |> count(STATUS, name = "Samples"),
               caption = "PLINK --check-sex status counts.")
    prob <- sxt |> filter(!str_detect(STATUS, regex("^ok$", ignore_case = TRUE)))
    if (nrow(prob)) {
      cat("**", n_fmt(nrow(prob)), " sample(s) were discordant and removed.**", sep = "")
      if (have(sx_prob)) cat(" The FID/IID list the workflow acted on is ",
                             fref(sx_prob), ".", sep = "")
      cat("\n\n")
      ## dont show the ids if discordant samples 
      # show_table(head(prob, 50))
      # if (nrow(prob) > 50) cat("_(showing the first 50 of ", nrow(prob), ")_\n\n", sep = "")
    } else {
      cat("No sex discordances were found.\n\n")
    }
  } else {
    show_table(head(sxt, 20), caption = "Head of the sex check report.")
  }

  if (have(sx_kept)) {
    cat("The dataset written by `call-RemoveSexFails` holds ", n_fmt(nlines(sx_kept)),
        " samples: ", fref(sx_kept), ".\n\n", sep = "")
  }
}
```

# Ancestry (step 4)

Principal components are computed on the **1000 Genomes reference panel alone**,
and the study samples are then projected onto those axes with PLINK 2
`--score`. Because no study sample enters the eigendecomposition, related or
over-represented individuals cannot bend the axes — the coordinate system is
fixed by the reference and identical between runs. A random forest trained on
the labelled reference samples assigns each study sample a superpopulation and
a probability; samples below the configured probability threshold are labelled
`unassigned` rather than forced into a cluster.

These labels drive the ancestry-aware filtering that follows: HWE is tested
within ancestry, heterozygosity is judged within ancestry, and the step 13
subset is defined by them.

```{r ancestry-plot}
#| results: asis

p <- pick("plots", "_ancestry_pca\\.png$")
if (have(p)) {
  show_image(p, "Study samples projected onto 1000 Genomes principal components")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No ancestry PCA plot under `call-AncestryPCA/`.")
}
```

```{r ancestry-counts}
#| results: asis
#| fig-height: 3.2

if (is.null(anc_tbl)) {
  absent("No `*_ancestry_assignments.tsv` — ancestry summaries are unavailable.")
} else {
  cat("Source file: ", fref(anc_file), " (columns: ",
      paste0("`", names(anc_tbl), "`", collapse = ", "), ")\n\n", sep = "")
}
```

```{r ancestry-counts-plot}
#| fig-height: 3.2

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  ac <- anc_tbl |> count(superpop, name = "n") |> arrange(desc(n))
  ggplot(ac, aes(x = reorder(superpop, n), y = n)) +
    geom_col(fill = PAL[["blue"]], width = 0.7) +
    geom_text(aes(label = scales::comma(n)), hjust = -0.2,
              size = 3, colour = PAL[["ink2"]]) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Assigned superpopulation",
         subtitle = "Counts are labelled directly; ancestry is on the axis, not in the colour",
         x = NULL, y = "Samples")
}
```

```{r ancestry-tables}
#| results: asis
#| fig-height: 3

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  ac <- anc_tbl |>
    count(superpop, name = "Samples") |>
    mutate(`% of cohort` = p_fmt(Samples / sum(Samples))) |>
    arrange(desc(Samples)) |>
    rename(Superpopulation = superpop)
  show_table(ac, caption = "Superpopulation assignment counts.")

  if ("probability" %in% names(anc_tbl)) {
    un <- sum(anc_tbl$superpop == "unassigned")
    if (un > 0) {
      note(sprintf(paste("%s sample(s) fell below the assignment probability",
                         "threshold and are labelled `unassigned`. They stay in the",
                         "dataset — they are simply excluded from ancestry-specific",
                         "tests and from the step 13 subset."), n_fmt(un)))
    }
  }
}
```

```{r ancestry-prob-plot}
#| fig-height: 3

if (!is.null(anc_tbl) && all(c("probability", "superpop") %in% names(anc_tbl))) {
  ggplot(anc_tbl, aes(x = probability)) +
    geom_histogram(bins = 40, fill = PAL[["blue"]], colour = "white", linewidth = 0.2) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
    labs(title = "Random forest assignment confidence",
         subtitle = "A mass near 1.0 means most samples sit cleanly inside a reference cluster",
         x = "Assignment probability", y = "Samples")
}
```

```{r ancestry-scree}
#| results: asis
#| fig-height: 3

ev_all <- picks("pca", "\\.eigenval$")
ev <- ev_all[!grepl("covariate", basename(ev_all))]
ev <- if (length(ev)) ev[1] else NA_character_
if (have(ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    scree <- tibble(PC = seq_along(vals), eigenvalue = vals,
                    pct = vals / sum(vals))
    cat("Variance carried by each reference PC (source: ", fref(ev), "):\n\n", sep = "")
    show_table(
      scree |> mutate(`Variance explained` = p_fmt(pct)) |>
        select(PC, Eigenvalue = eigenvalue, `Variance explained`) |>
        mutate(Eigenvalue = round(Eigenvalue, 3)),
      caption = "Ancestry PCA eigenvalues."
    )
  }
}
```

# Minor allele count filter (step 5)

Rare variants are removed on the full cohort, since rarity is a property of the
whole sample, not of one ancestry. Monomorphic sites go first, then everything
below the configured MAC threshold. Low-count variants carry almost no
association information and are the main source of spurious HWE and imputation
failures.

```{r mac}
#| results: asis

p <- pick("plots", "mac_plot\\.png$")
if (have(p)) {
  show_image(p, "Minor allele count distribution before and after filtering")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No `mac_plot.png` under `call-MacFilter/`.")
}

fb <- pick("reports", "^freq_before\\.frq$")
fa <- pick("reports", "^freq_after\\.frq$")
if (have(fb) || have(fa)) {
  ## These files carry one row per variant (tens of MB in a real run), so only
  ## the MAF column is parsed.
  summarise_frq <- function(p, label) {
    if (!have(p)) return(NULL)
    d <- suppressWarnings(read_table(p, col_types = cols_only(MAF = col_double()),
                                     progress = FALSE))
    if (!"MAF" %in% names(d)) return(NULL)
    tibble(
      Stage          = label,
      Variants       = nrow(d),
      `Monomorphic`  = sum(d$MAF == 0, na.rm = TRUE),
      `MAF < 1%`     = sum(d$MAF < 0.01, na.rm = TRUE),
      `MAF < 5%`     = sum(d$MAF < 0.05, na.rm = TRUE),
      `Median MAF`   = round(median(d$MAF, na.rm = TRUE), 4)
    )
  }
  frq <- bind_rows(summarise_frq(fb, "Before MAC filter"),
                   summarise_frq(fa, "After MAC filter"))
  if (!is.null(frq) && nrow(frq)) {
    show_table(frq, caption = paste0("Allele frequency summary (sources: ",
                                     rel(fb), ", ", rel(fa), ")."))
  }
} else {
  absent("No `freq_before.frq` / `freq_after.frq` under `call-MacFilter/`.")
}
```

# Hardy–Weinberg equilibrium (step 6)

HWE violations are **detected within each tested ancestry** and then removed
from the full cohort. Testing the pooled cohort would flag variants that are
perfectly in equilibrium within each population and out of it only because the
cohort mixes populations — the Wahlund effect. Detecting per ancestry and
removing globally keeps one variant set across the whole dataset while avoiding
that artefact.

```{r hwe}
#| results: asis

hwe_line <- log_txt[str_detect(log_txt, regex("^Step 6\\s+HWE", ignore_case = TRUE))]
mac_line <- log_txt[str_detect(log_txt, regex("^Step 5", ignore_case = TRUE))]

if (length(hwe_line) && length(mac_line)) {
  n_before <- as.integer(str_match(mac_line[1], "SNPs:\\s*(\\d+)")[, 2])
  n_after  <- as.integer(str_match(hwe_line[1], "SNPs:\\s*(\\d+)")[, 2])
  if (!is.na(n_before) && !is.na(n_after)) {
    show_table(
      tibble(Metric = c("Variants before HWE filter", "Variants after HWE filter",
                        "Removed for HWE violation"),
             Value = n_fmt(c(n_before, n_after, n_before - n_after))),
      caption = "HWE filtering, from the pipeline log.", align = c("l", "r")
    )
  }
} else {
  absent("The pipeline log has no step 5/6 count lines to compare.")
}

## The run directory keeps the actual list of removed variants, which the
## collected results do not.
hwe_removed <- NA_character_
d_hwe <- exec_dir("HweRemoved")
if (!is.na(d_hwe)) {
  hits <- list.files(d_hwe, pattern = "removed_snps\\.txt$", full.names = TRUE)
  if (length(hits)) hwe_removed <- hits[1]
}
if (have(hwe_removed)) {
  cat("The ", n_fmt(nlines(hwe_removed)), " variant IDs removed for HWE violation ",
      "are listed in ", fref(hwe_removed), ".\n\n", sep = "")
      # " — this list is not copied out by `collect_cromwell_results.bash`.\n\n", sep = "")
}

# hwe_files <- picks("reports", "\\.hwe$")
# if (length(hwe_files)) {
#   cat("Per-variant HWE statistics: ", paste(sprintf("`%s`", rel(hwe_files)),
#       collapse = ", "), "\n\n", sep = "")
# } else {
#   note(paste("There is no `.hwe` table: the workflow applies `--hwe` as a",
#              "filter rather than running `--hardy`, so PLINK writes no per-variant",
#              "HWE report. The counts above are the authoritative record of what",
#              "the step removed."))
# }
```

# Heterozygosity (steps 7a–7b)

Step 7a LD-prunes the dataset with high-LD regions excluded. The pruned SNP
list is a **helper only** — it feeds the heterozygosity and relatedness checks
and is never applied to the delivered genotypes.

Step 7b flags samples more than 3 SD from the mean heterozygosity rate.
Excess heterozygosity suggests contamination or a mixed sample; a deficit
suggests inbreeding or poor DNA quality. Detection and removal both happen
within the ancestry subset, so samples from untested populations are not judged
against a mean that does not describe them.

```{r het}
#| results: asis

p <- pick("plots", "heterozygosity\\.png$")
if (have(p)) {
  show_image(p, "Heterozygosity rate distribution with the ±3 SD outlier bounds")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No `heterozygosity.png` under `call-HeterozygosityCheck/`.")
}

pr <- pick("", "\\.prune\\.in$")
if (!have(pr)) {
  d <- exec_dir("LdPruningHet")
  if (!is.na(d)) {
    hits <- list.files(d, pattern = "\\.prune\\.in$", full.names = TRUE)
    if (length(hits)) pr <- hits[1]
  }
}
if (have(pr)) {
  cat("Step 7a kept ", n_fmt(nlines(pr)), " independent SNPs: ", fref(pr),
      ". The same list is reused by the relatedness check at step 9.\n\n", sep = "")
}

het <- pick("reports", "R_check\\.het$")
hf  <- pick("reports", "het_fail_samples\\.txt$")

if (have(het)) {
  h <- suppressWarnings(read_table(het, show_col_types = FALSE, progress = FALSE))
  nm <- names(h)
  if (all(c("N(NM)", "O(HOM)") %in% nm) || all(c("N.NM.", "O.HOM.") %in% nm)) {
    n_nm  <- if ("N(NM)" %in% nm) h[["N(NM)"]] else h[["N.NM."]]
    o_hom <- if ("O(HOM)" %in% nm) h[["O(HOM)"]] else h[["O.HOM."]]
    rate  <- (n_nm - o_hom) / n_nm
    show_table(
      tibble(Metric = c("Samples tested", "Mean heterozygosity rate",
                        "SD", "Lower bound (mean − 3 SD)", "Upper bound (mean + 3 SD)"),
             Value  = c(n_fmt(length(rate)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE)),
                        sprintf("%.4f", sd(rate, na.rm = TRUE)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE) - 3 * sd(rate, na.rm = TRUE)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE) + 3 * sd(rate, na.rm = TRUE)))),
      caption = paste0("Heterozygosity summary (source: ", rel(het), ")."),
      align = c("l", "r")
    )
  }
} else {
  absent("No `R_check.het` under `call-HeterozygosityCheck/`.")
}

if (have(hf)) {
  n_fail <- nrow(read_id_file(hf))
  cat("**", n_fmt(n_fail), " sample(s) failed the heterozygosity check** and were removed. ",
      "Full list: ", fref(hf), "\n\n", sep = "")
  ## don't show the list of failed samples
  # if (n_fail > 0 && n_fail <= 50) {
  #   ff <- read_id_file(hf)
  #   show_table(ff, caption = "Samples removed as heterozygosity outliers.")
  # }
} else {
  absent("No `het_fail_ind.txt` under `call-HeterozygosityCheck/`.")
}
```

# Chromosome filter (step 8)

The dataset is restricted to the configured chromosome range (`--chr 1-23` by
default). This is the **final QC dataset**: everything after it is
harmonisation and packaging, not filtering on QC grounds.

```{r chrfilter}
#| results: asis

l8 <- step_counts |> filter(str_detect(step, regex("^Step 8", ignore_case = TRUE)))
if (nrow(l8)) {
  show_table(l8 |> select(Step = step, Variants = snps, Samples = samples),
             caption = "Final QC dataset (step 8).")
} else {
  absent("No step 8 line in the pipeline log.")
}

## Unlike the collected results, the run directory still holds this dataset.
d8 <- exec_dir("ChromosomeFilter")
if (!is.na(d8)) {
  b8 <- list.files(d8, pattern = "\\.bim$", full.names = TRUE)
  f8 <- list.files(d8, pattern = "\\.fam$", full.names = TRUE)
  if (length(b8) && length(f8)) {
    cat("The step 8 fileset itself is here — ", n_fmt(nlines(b8[1])), " variants × ",
        n_fmt(nlines(f8[1])), " samples in `", rel(sub("\\.bim$", "", b8[1])),
        ".{bed,bim,fam}`.\n\n", sep = "")
    note(paste("`collect_cromwell_results.bash` deliberately does **not** copy this",
               "fileset out. What it puts in `qc_dataset/` is the step 10 harmonised",
               "dataset, so that the PLINK files and the imputation VCFs describe",
               "exactly the same variants. Use the step 8 data only if you",
               "specifically want the pre-harmonisation genotypes."))
  }
}
```

# Relatedness (step 9)

KING-robust kinship is computed on the final QC dataset, reusing the step 7a
pruned SNP list. Running it here rather than earlier is deliberate: the
ancestry PCA takes its axes from the reference panel alone, so relatives could
not have distorted it, and computing kinship on the post-MAC, post-HWE variant
set gives a cleaner estimate.

**Related samples are flagged, not removed.** Which member of a pair to drop
depends on the analysis — phenotype completeness, case status, genotyping
quality — so the pipeline leaves the decision to you and ships both lists.

```{r related}
#| results: asis
#| fig-height: 3.2

rout <- pick("reports", "relatedness_flagged_samples\\.txt$")
rin  <- pick("reports", "relatedness_unrelated_samples\\.txt$")
gen  <- pick("reports", "\\.genome$")

rows <- list()
if (have(rout)) rows[[length(rows) + 1]] <-
  tibble(Metric = "Samples flagged as related (to remove for an unrelated set)",
         Value = n_fmt(nrow(read_id_file(rout))), File = rel(rout))
if (have(rin)) rows[[length(rows) + 1]] <-
  tibble(Metric = "Samples in the maximal unrelated set",
         Value = n_fmt(nrow(read_id_file(rin))), File = rel(rin))

gtab <- NULL
if (have(gen)) {
  gtab <- suppressWarnings(read_table(gen, show_col_types = FALSE, progress = FALSE))
  rows[[length(rows) + 1]] <-
    tibble(Metric = "Sample pairs above the pi-hat reporting threshold",
           Value = n_fmt(nrow(gtab)), File = rel(gen))
}

if (length(rows)) {
  show_table(bind_rows(rows), align = c("l", "r", "l"))
} else {
  absent("No relatedness outputs under `call-RelatednessCheck/`.")
}
```

```{r related-plot}
#| fig-height: 3.2

if (!is.null(gtab) && "PI_HAT" %in% names(gtab) && nrow(gtab) > 0) {
  ggplot(gtab, aes(x = PI_HAT)) +
    geom_histogram(bins = 40, fill = PAL[["blue"]], colour = "white", linewidth = 0.2) +
    geom_vline(xintercept = 0.5,   colour = PAL[["orange"]], linewidth = 0.6, linetype = "dashed") +
    geom_vline(xintercept = 0.25,  colour = PAL[["orange"]], linewidth = 0.6, linetype = "dashed") +
    annotate("text", x = 0.5,  y = Inf, label = "1st degree (0.5)", vjust = 1.6,
             hjust = -0.05, size = 2.8, colour = PAL[["orange"]]) +
    annotate("text", x = 0.25, y = Inf, label = "2nd degree (0.25)", vjust = 3.2,
             hjust = -0.05, size = 2.8, colour = PAL[["orange"]]) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
    labs(title = "Pi-hat distribution of reported pairs",
         subtitle = "Only pairs above the configured pihat_min appear in the .genome file",
         x = "PI_HAT (proportion IBD)", y = "Pairs")
}
```

```{r related-pairs}
#| results: asis
#| eval: false

## dont show the table of failed samples
# if (!is.null(gtab) && nrow(gtab) > 0) {
#   cols <- intersect(c("FID1", "IID1", "FID2", "IID2", "Z0", "Z1", "Z2", "PI_HAT"),
#                     names(gtab))
#   top <- gtab |> arrange(desc(.data[["PI_HAT"]])) |> head(25) |> select(all_of(cols))
#   show_table(top, caption = "The 25 most closely related pairs (full table in the .genome file).")
#   if (nrow(gtab) > 25) cat("_(", n_fmt(nrow(gtab)), " pairs in total)_\n\n", sep = "")
# }
```

# Imputation preparation (step 10)

Will Rayner's `HRC-1000G-check-bim.pl` harmonises the QC dataset against the
imputation reference panel: it fixes strand flips, aligns ref/alt, and drops
variants the panel cannot accept — sites absent from the panel, palindromic
A/T and C/G SNPs that cannot be strand-resolved, allele mismatches, and
variants whose frequency differs from the reference by more than 0.2.

The step emits one VCF per chromosome for upload **and** merges the harmonised
per-chromosome filesets back into a single `bed/bim/fam`. That combined file is
the PLINK equivalent of the VCFs — same variants, same alleles — which is what
makes the two deliverables directly comparable.

```{r checkbim}
#| results: asis

cb <- pick("reports", "check-bim\\.log$")
if (have(cb)) {
  txt <- read_lines(cb, progress = FALSE)
  m <- str_match(txt, "^\\s*([A-Za-z][^/]*?)\\s+(\\d+)\\s*$")
  ok <- !is.na(m[, 1]) & nchar(str_trim(m[, 2])) > 3
  if (any(ok)) {
    show_table(
      tibble(Check = str_trim(m[ok, 2]), Variants = as.integer(m[ok, 3])),
      caption = paste0("Counts parsed from ", rel(cb), ".")
    )
  }
  collapsible("check-bim.log (full)", txt)
} else {
  absent("No `check-bim.log` under `call-PrepareForImputation/`.")
}

## The exclusion lists check-bim writes live only in the run directory.
dpi <- exec_dir("PrepareForImputation")
if (!is.na(dpi)) {
  lists <- list.files(dpi, full.names = TRUE,
                      pattern = "^(Exclude|Strand-Flip|Force-Allele1|Position|Chromosome|ID)-.*\\.txt$")
  if (length(lists)) {
    li <- tibble(File = basename(lists),
                 Variants = vapply(lists, nlines, integer(1))) |> arrange(File)
    show_table(li, caption = paste0(
      "Per-action variant lists written by check-bim in `", rel(dpi),
      "`. These are not copied out by `collect_cromwell_results.bash`."))
  }
}
```

## Variants per chromosome in the imputation-ready data

```{r varchr-final}
#| results: asis
#| fig-height: 3.8

vc_fin <- read_varchr(pick("reports", "final_variants_per_chr\\.txt$"))
if (is.null(vc_fin)) {
  absent("No `variants_per_chr_report.txt` under `call-PostImputationVariantsPerChromosome/`.")
} else {
  cat("These counts match the VCFs variant-for-variant. Source file: ",
      fref(pick("reports", "final_variants_per_chr\\.txt$")), "\n\n", sep = "")
}
```

```{r varchr-final-plot}
#| fig-height: 3.8

if (!is.null(vc_fin)) {
  if (!is.null(vc_init)) {
    cmp <- bind_rows(
      vc_init |> mutate(stage = "Entering QC"),
      vc_fin  |> mutate(stage = "Imputation-ready")
    ) |>
      mutate(stage = factor(stage, levels = c("Entering QC", "Imputation-ready")),
             chr_label = factor(chr_label, levels = unique(vc_init$chr_label)))

    ggplot(cmp, aes(x = chr_label, y = variants, fill = stage)) +
      geom_col(position = position_dodge(width = 0.78), width = 0.68) +
      scale_fill_manual(values = unname(PAL[c("blue", "orange")])) +
      scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
      labs(title = "Variants per chromosome, before QC and after imputation prep",
           subtitle = "The gap is what missingness, MAC, HWE and panel harmonisation removed",
           x = "Chromosome", y = "Variants")
  } else {
    ggplot(vc_fin, aes(x = factor(chr_label, levels = chr_label), y = variants)) +
      geom_col(fill = PAL[["blue"]], width = 0.72) +
      scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
      labs(title = "Variants per chromosome in the imputation-ready data",
           x = "Chromosome", y = "Variants")
  }
}
```

```{r varchr-final-table}
#| results: asis

if (!is.null(vc_fin)) {
  tab <- vc_fin |> select(Chromosome = chr_label, `Imputation-ready` = variants)
  if (!is.null(vc_init)) {
    tab <- vc_init |>
      select(Chromosome = chr_label, `Entering QC` = variants) |>
      full_join(tab, by = "Chromosome") |>
      mutate(Removed = `Entering QC` - `Imputation-ready`,
             `Retained` = p_fmt(`Imputation-ready` / `Entering QC`))
  }
  show_table(tab, caption = "Per-chromosome variant counts.")
}
```

## Imputation VCFs

```{r vcfs}
#| results: asis

vcf_files <- picks("vcfs", "\\.vcf\\.gz$")
tbi_files <- picks("vcfs", "\\.vcf\\.gz\\.tbi$")

if (length(vcf_files) == 0) {
  absent("No VCFs under `call-PrepareForImputation/` — the imputation deliverable is missing.")
} else {
  vi <- tibble(File = basename(vcf_files), Bytes = file.info(vcf_files)$size) |>
    mutate(chr = as.integer(str_match(File, "chr(\\d+)\\.vcf\\.gz$")[, 2]),
           Size = hsize(Bytes),
           Index = ifelse(paste0(File, ".tbi") %in% basename(tbi_files), "yes", "MISSING")) |>
    arrange(chr) |>
    select(File, Size, `Tabix index` = Index)
  show_table(vi, caption = sprintf(
    "Imputation-ready VCFs in `%s` (%d files, %s total).",
    rel(dirname(vcf_files[1])), length(vcf_files),
    hsize(sum(file.info(vcf_files)$size))))

  if (any(vi$`Tabix index` == "MISSING")) {
    absent("One or more VCFs have no `.tbi` index — imputation servers will reject them.")
  } else {
    note(paste("Upload these to the imputation server (HRC or TOPMed, matching the",
               "panel the run was harmonised against). Run",
               "`collect_cromwell_results.bash` to gather them into a `vcfs/`",
               "directory first — the run directory also holds a duplicate copy of",
               "each VCF under `execution/glob-*/`."))
  }
}
```

# Combined imputation-ready dataset (step 11)

The same harmonised data in PLINK form, in both PLINK 1 and PLINK 2 encodings.
Use these for any analysis you want to run on the pre-imputation genotypes;
they are variant-for-variant identical to the uploaded VCFs.

```{r combined}
#| results: asis

fam <- pick("qc_dataset", "\\.fam$")
bim <- pick("qc_dataset", "\\.bim$")
bed <- pick("qc_dataset", "\\.bed$")
pgen <- pick("qc_dataset", "\\.pgen$")
pvar <- pick("qc_dataset", "\\.pvar$")
psam <- pick("qc_dataset", "\\.psam$")

if (!have(bim) && !have(pgen)) {
  absent("No combined dataset under `call-PrepareForImputation/` or `call-CombinedToPlink2/`.")
} else {
  rows <- list()
  if (have(bim)) rows[[length(rows) + 1]] <-
    tibble(Format = "PLINK 1 (bed/bim/fam)",
           Variants = n_fmt(nlines(bim)), Samples = n_fmt(nlines(fam)),
           Files = paste(sprintf("`%s`", basename(c(bed, bim, fam))), collapse = ", "))
  if (have(psam)) rows[[length(rows) + 1]] <-
    tibble(Format = "PLINK 2 (pgen/pvar/psam)",
           Variants = if (have(pvar)) n_fmt(sum(!str_starts(
                        read_lines(pvar, progress = FALSE), "#"))) else "—",
           Samples = n_fmt(max(0, nlines(psam) - 1)),
           Files = paste(sprintf("`%s`", basename(c(pgen, pvar, psam))), collapse = ", "))
  show_table(bind_rows(rows), caption = paste0(
    "The combined dataset, in `call-PrepareForImputation/` and `call-CombinedToPlink2/`."))
}
```

# Covariate PCs (step 12)

Within-cohort principal components for use as association covariates. The axes
are estimated on the KING-unrelated samples — relatives would otherwise pull
components towards their own family structure — and then **all** samples are
projected onto them, so nobody is dropped from the covariate file.

These are a different object from the step 4 ancestry PCs: those place samples
against an external reference to label ancestry, these describe residual
structure inside this cohort.

```{r covar}
#| results: asis
#| fig-height: 3

cov_f  <- pick("pca", "_covariate_pca_covariates\\.tsv$")
if (!have(cov_f)) cov_f <- pick("pca", "_covariates\\.tsv$")
cov_ev <- pick("pca", "_covariate_pca\\.eigenval$")

cov_tbl <- if (have(cov_f)) suppressWarnings(read_tsv(cov_f, show_col_types = FALSE,
                                                      progress = FALSE)) else NULL

if (is.null(cov_tbl)) {
  absent("No covariate PC file under `call-CovariatePCA/`.")
} else {
  npc <- length(pc_cols(cov_tbl))
  cat("**", n_fmt(nrow(cov_tbl)), " samples × ", npc, " PCs** — ", fref(cov_f),
      ". Columns: `FID`, `IID`, `PC1`…`PC", npc, "`.\n\n", sep = "")
}
```

```{r covar-scree}
#| fig-height: 3

scree <- NULL
if (have(cov_ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(cov_ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    scree <- tibble(PC = seq_along(vals), eigenvalue = vals, pct = vals / sum(vals))
    ggplot(scree, aes(x = PC, y = pct)) +
      geom_col(fill = PAL[["blue"]], width = 0.7) +
      geom_text(data = head(scree, 5), aes(label = p_fmt(pct)),
                vjust = -0.6, size = 2.8, colour = PAL[["ink2"]]) +
      scale_x_continuous(breaks = scales::pretty_breaks()) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
      labs(title = "Variance explained by each within-cohort PC",
           subtitle = "Where the bars flatten is a reasonable place to stop taking covariates",
           x = "Principal component", y = "Share of variance")
  }
}
```

```{r covar-scree-table}
#| results: asis

if (!is.null(scree)) {
  show_table(
    scree |> head(10) |>
      mutate(Eigenvalue = round(eigenvalue, 3), `Variance explained` = p_fmt(pct)) |>
      select(PC, Eigenvalue, `Variance explained`),
    caption = paste0("First 10 within-cohort eigenvalues (source: ", rel(cov_ev), ").")
  )
}
```

```{r covar-scatter}
#| fig-height: 5.5

## PC1/PC2 as small multiples by ancestry: identity comes from the facet strip,
## so the chart needs no categorical palette and stays readable under CVD.
if (!is.null(qc_tbl) && all(c("PC1", "PC2", "ancestry") %in% names(qc_tbl))) {
  d <- qc_tbl |> filter(!is.na(PC1), !is.na(PC2))
  if (nrow(d) > 0 && n_distinct(d$ancestry) > 1) {
    ggplot(d, aes(x = PC1, y = PC2)) +
      geom_point(data = transform(d, ancestry = NULL), colour = PAL[["grey"]],
                 alpha = 0.35, size = 0.5) +
      geom_point(colour = PAL[["blue"]], alpha = 0.75, size = 0.7) +
      facet_wrap(~ancestry) +
      labs(title = "Within-cohort PC1 vs PC2, by assigned ancestry",
           subtitle = "Grey shows the whole cohort in every panel for context",
           x = "PC1", y = "PC2")
  } else {
    ggplot(d, aes(x = PC1, y = PC2)) +
      geom_point(colour = PAL[["blue"]], alpha = 0.7, size = 0.8) +
      labs(title = "Within-cohort PC1 vs PC2", x = "PC1", y = "PC2")
  }
}
```

# Sample QC status table

One row per participant, joining ancestry, the relatedness flag and the
covariate PCs. This single file is enough to subset the cohort and pull
covariates without touching the genotype files — for example, keep
`ancestry == "EUR" & !related`, then pass `PC1..PC10` to your association model.

```{r qcstatus}
#| results: asis

if (is.null(qc_tbl)) {
  absent("No `*_sample_qc_status.tsv` under `call-CreateQCStatusTable/`.")
} else {
  cat("Source file: ", fref(qc_file), " — ", n_fmt(nrow(qc_tbl)), " samples, ",
      "columns: ", paste0("`", names(qc_tbl)[seq_len(min(6, ncol(qc_tbl)))], "`",
      collapse = ", "),
      if (ncol(qc_tbl) > 6) sprintf(", … (%d columns in total)", ncol(qc_tbl)) else "",
      "\n\n", sep = "")

  ## The table is built from the step 4 ancestry assignments, which were made
  ## before the heterozygosity filter. If samples were dropped after step 4 the
  ## table lists more people than the delivered genotypes contain.
  n_final <- if (have(fam)) nlines(fam) else NA_integer_
  if (!is.na(n_final) && nrow(qc_tbl) != n_final) {
    note(sprintf(paste("This table has %s rows but the delivered dataset holds %s",
                       "samples. The extra rows are samples assigned an ancestry at",
                       "step 4 and dropped later (heterozygosity outliers, and sex",
                       "discordances if the check ran). Join on `FID`/`IID` against",
                       "the `.fam`/`.psam` before using it as a cohort list."),
                 n_fmt(nrow(qc_tbl)), n_fmt(n_final)))
  }

  if (all(c("ancestry", "related") %in% names(qc_tbl))) {
    ct <- qc_tbl |>
      mutate(related = ifelse(as.logical(related), "Related (flagged)", "Unrelated")) |>
      count(ancestry, related) |>
      pivot_wider(names_from = related, values_from = n, values_fill = 0)
    if (!"Related (flagged)" %in% names(ct)) ct[["Related (flagged)"]] <- 0L
    if (!"Unrelated" %in% names(ct)) ct[["Unrelated"]] <- 0L
    ct <- ct |>
      mutate(Total = Unrelated + `Related (flagged)`) |>
      arrange(desc(Total)) |>
      rename(Ancestry = ancestry)
    ct <- bind_rows(ct, summarise(ct, Ancestry = "All",
                                  across(where(is.numeric), sum)))
    show_table(ct, caption = "Cohort composition by ancestry and relatedness flag.")
  }

  ## Don't show the samples

  ## Show the leading columns only — with 20+ PCs the full table does not fit
  ## on a page, and the point here is the shape of the file, not its contents.
  # head_cols <- intersect(c("FID", "IID", "ancestry", "ancestry_prob", "related",
  #                          "PC1", "PC2", "PC3"), names(qc_tbl))
  # show_table(head(qc_tbl, 5) |> select(all_of(head_cols)) |>
  #              mutate(across(where(is.numeric), \(x) round(x, 4))),
  #            caption = paste0("First five rows of the sample QC status table",
  #                             if (ncol(qc_tbl) > length(head_cols))
  #                               sprintf(" (first %d of %d columns).",
  #                                       length(head_cols), ncol(qc_tbl)) else "."))
}
```

# Unrelated single-ancestry subset (step 13)

An additional deliverable: the imputation-ready data restricted to samples of
one superpopulation that are not KING-flagged as related, with covariate PCs
recomputed **inside** the subset. Whole-cohort PCs largely capture
between-ancestry structure, which is not the axis you want to adjust for in a
within-population analysis — hence the separate PCA.

```{r subset}
#| results: asis
#| fig-height: 3

s_bim  <- pick("subset", "\\.bim$")
s_fam  <- pick("subset", "\\.fam$")
s_bed  <- pick("subset", "\\.bed$")
s_pgen <- pick("subset", "\\.pgen$")
s_pvar <- pick("subset", "\\.pvar$")
s_psam <- pick("subset", "\\.psam$")
s_keep <- pick("subset", "_keep\\.txt$")
s_cov  <- pick("subset", "_covariates\\.tsv$")
s_ev   <- pick("subset", "\\.eigenval$")

subset_pop <- if (have(s_keep)) str_match(basename(s_keep),
                  "_unrelated_([A-Za-z]+)_keep\\.txt$")[, 2] else NA_character_

if (!have(s_bim) && !have(s_keep)) {
  absent("Nothing under `call-SubsetUnrelated/` — step 13 did not run.")
} else {
  if (!is.na(subset_pop)) {
    cat("Subset population: **", subset_pop, "**, unrelated samples only.\n\n", sep = "")
  }
  rows <- list()
  if (have(s_keep)) rows[[length(rows) + 1]] <-
    tibble(Item = "Samples in the subset", Value = n_fmt(nlines(s_keep)), File = rel(s_keep))
  if (have(s_bim)) rows[[length(rows) + 1]] <-
    tibble(Item = "Variants (PLINK 1)", Value = n_fmt(nlines(s_bim)),
           File = paste(basename(c(s_bed, s_bim, s_fam)), collapse = ", "))
  if (have(s_psam)) rows[[length(rows) + 1]] <-
    tibble(Item = "PLINK 2 copy", Value = "pgen/pvar/psam",
           File = paste(basename(c(s_pgen, s_pvar, s_psam)), collapse = ", "))
  if (have(s_cov)) {
    sc <- suppressWarnings(read_tsv(s_cov, show_col_types = FALSE, progress = FALSE))
    rows[[length(rows) + 1]] <-
      tibble(Item = "Within-subset covariate PCs",
             Value = sprintf("%s samples × %d PCs", n_fmt(nrow(sc)), length(pc_cols(sc))),
             File = rel(s_cov))
  }
  show_table(bind_rows(rows), align = c("l", "r", "l"),
             caption = "Step 13 outputs, across the three tasks that build the subset.")
}
```

```{r subset-scree}
#| fig-height: 3

if (have(s_ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(s_ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    sc <- tibble(PC = seq_along(vals), pct = vals / sum(vals))
    ggplot(sc, aes(x = PC, y = pct)) +
      geom_col(fill = PAL[["blue"]], width = 0.7) +
      geom_text(data = head(sc, 5), aes(label = p_fmt(pct)),
                vjust = -0.6, size = 2.8, colour = PAL[["ink2"]]) +
      scale_x_continuous(breaks = scales::pretty_breaks()) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
      labs(title = paste0("Variance explained within the ",
                          ifelse(is.na(subset_pop), "", paste0(subset_pop, " ")),
                          "unrelated subset"),
           subtitle = "Flatter than the whole-cohort scree — between-ancestry structure is gone",
           x = "Principal component", y = "Share of variance")
  }
}
```

# What to use for what

```{r deliverables}
#| results: asis

d <- tribble(
  ~Purpose, ~`Task directory (under the run directory)`,
  "Upload to the imputation server (HRC / TOPMed)", "`call-PrepareForImputation/execution/*_chr*.vcf.gz` + `.tbi`",
  "Pre-imputation analysis in PLINK 1", "`call-PrepareForImputation/execution/*_combined.{bed,bim,fam}`",
  "Pre-imputation analysis in PLINK 2", "`call-CombinedToPlink2/execution/*.{pgen,pvar,psam}`",
  "Association covariates for the whole cohort", "`call-CovariatePCA/execution/*_covariates.tsv`",
  "Subsetting the cohort by ancestry or relatedness", "`call-CreateQCStatusTable/execution/*_sample_qc_status.tsv`",
  "Dropping relatives from an analysis", "`call-RelatednessCheck/execution/plink2.king.cutoff.out.id` with PLINK `--remove`",
  "A ready-made unrelated single-ancestry analysis set", "`call-SubsetUnrelated/`, `call-SubsetToPlink2/`, `call-SubsetCovariatePCA/`",
  "Per-step counts for a methods section", "`call-WriteLog/execution/*_pipeline.log`",
  "The pre-harmonisation step 8 dataset", "`call-ChromosomeFilter/execution/*_QC8.{bed,bim,fam}`"
)
show_table(d)

note(paste("Run `bash src/collect_cromwell_results.bash <run-dir> ./results` before",
           "handing any of this on. The run directory mixes deliverables with every",
           "intermediate dataset, duplicate copies under `execution/glob-*/` and",
           "`inputs/`, and Cromwell's own scaffolding — it is not a distributable",
           "result set."))

note(paste("Related samples are **flagged, not removed**, everywhere except the",
           "step 13 subset. If your analysis assumes unrelated individuals, apply",
           "`--remove` with the KING cutoff list yourself."))
```

# Task inventory

Every task in the run, with its exit code and what it produced. Cromwell
scaffolding (`script`, `stdout`, `stderr`, `rc`),
PLINK side files (`.hh`, `.nosex`, `.log`) and duplicate copies under
`glob-*/` and `inputs/` are excluded from the file counts.

```{r inventory}
#| results: asis

SCAFFOLD <- paste0("^(script|script\\.submit|script\\.background|stdout|stderr|",
                   "stdout\\.background|stderr\\.background|rc|tmp\\..*|Rplots\\.pdf)$")

inv <- lapply(tasks, function(tk) {
  d <- exec_dir(tk)
  if (is.na(d)) return(NULL)
  f <- list.files(d, full.names = TRUE)
  f <- f[!dir.exists(f)]                                  # drops glob-*/ dirs
  f <- f[!str_detect(basename(f), SCAFFOLD)]
  ## PLINK's own .log/.hh/.nosex side files are noise in this column, but the
  ## pipeline log and the check-bim log are real outputs.
  side <- str_detect(basename(f), "\\.(hh|nosex)$") |
          (str_detect(basename(f), "\\.log$") &
           !str_detect(basename(f), "(_pipeline\\.log|check-bim\\.log)$"))
  keep <- f[!side]
  info <- file.info(f)
  rcv <- task_status$rc[match(tk, task_status$task)]
  tibble(
    Task    = tk,
    Exit    = ifelse(is.na(rcv), "—", rcv),
    Files   = length(f),
    Size    = hsize(sum(info$size, na.rm = TRUE)),
    Outputs = if (length(keep) == 0) "" else
      paste0(paste(basename(head(keep, 4)), collapse = ", "),
             if (length(keep) > 4) sprintf(", … (%d more)", length(keep) - 4) else "")
  )
}) |> bind_rows()

if (nrow(inv) == 0) {
  absent("No task directories could be read.")
} else {
  total_files <- sum(inv$Files)
  all_f <- unlist(lapply(tasks, function(tk) {
    d <- exec_dir(tk); if (is.na(d)) return(character(0))
    f <- list.files(d, full.names = TRUE); f[!dir.exists(f)]
  }))
  show_table(inv, caption = sprintf(
    "%d tasks, %s files, %s in `%s` (excluding duplicates under glob-*/ and inputs/).",
    nrow(inv), n_fmt(total_files),
    hsize(sum(file.info(all_f)$size, na.rm = TRUE)), RD))
}
```

```{r warnings}
#| results: asis

## Warnings in the per-task PLINK logs never reach the collected results.
warn_rows <- lapply(tasks, function(tk) {
  d <- exec_dir(tk)
  if (is.na(d)) return(NULL)
  logs <- list.files(d, pattern = "\\.log$", full.names = TRUE)
  if (!length(logs)) return(NULL)
  msgs <- unlist(lapply(logs, function(l) {
    x <- read_lines(l, progress = FALSE)
    x[str_detect(x, regex("^warning", ignore_case = TRUE))]
  }))
  if (!length(msgs)) return(NULL)
  tibble(Task = tk, Warnings = length(msgs),
         Example = str_trunc(str_squish(msgs[1]), 110))
}) |> bind_rows()

if (nrow(warn_rows)) {
  cat("## PLINK warnings\n\n")
  show_table(warn_rows, caption = paste0(
    "Warnings found in the per-task `.log` files. Most are routine (heterozygous ",
    "haploid genotypes on the X chromosome, for instance); this table is here ",
    "because these logs are not copied into the collected results."))
}
```

# Run parameters and provenance

```{r params}
#| results: asis

cfg_path <- resolve_path(params$config_json)
if (nzchar(cfg_path) && file.exists(cfg_path)) {
  if (requireNamespace("jsonlite", quietly = TRUE)) {
    cfg <- jsonlite::fromJSON(cfg_path)
    ct <- tibble(Parameter = sub("^genotype_qc_preimputation\\.", "", names(cfg)),
                 Value = vapply(cfg, \(x) paste(as.character(x), collapse = ", "), ""))
    ## Show the QC knobs up front; file and tool paths go in the fold.
    qc_params <- c("output_prefix", "geno_threshold", "mind_threshold",
                   "mac_threshold", "hwe_pvalue", "ld_window_kb", "ld_step",
                   "ld_r2", "pihat_min", "king_cutoff", "n_pcs",
                   "n_covariate_pcs", "chr_args", "pca_reference_populations",
                   "test_populations", "ancestry_prob_threshold",
                   "subset_population")
    thresholds <- ct |>
      filter(Parameter %in% qc_params) |>
      arrange(match(Parameter, qc_params))
    other <- ct |> filter(!Parameter %in% qc_params)
    show_table(thresholds, caption = paste0("QC parameters used for this run (",
                                            basename(cfg_path), ")."))
    collapsible("All inputs from the config file",
                sprintf("%-45s %s", other$Parameter, other$Value))
  } else {
    absent("The `jsonlite` package is not installed, so the config file could not be parsed.")
  }
} else {
  note(paste("No config file was passed. Re-render with",
             "`-P config_json=config/genotype_qc_preimputation_inputs.json`",
             "to document the thresholds this run used."))
}

## Every task's command line is preserved in the run directory, which is the
## most exact provenance record available for the run.
d_any <- exec_dir("WriteLog")
if (!is.na(d_any) && file.exists(file.path(d_any, "script"))) {
  cat("Each task's exact command line is preserved in `call-<Task>/execution/script`.\n\n")
}

cat("\n**Report generated:** ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    " · **R:** ", R.version.string,
    " · **Cromwell run:** `", RD, "`",
    if (nzchar(params$prefix)) paste0(" · `prefix=", params$prefix, "`") else "",
    "\n\n", sep = "")
```

```{r sessioninfo}
#| results: asis

collapsible("R session info",
            capture.output(sessionInfo()))
```
