---
title: "Genotype QC — pre-imputation report"
date: today
date-format: "YYYY-MM-DD"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-expand: 2
    number-sections: true
    embed-resources: true
    fig-align: center
    tbl-cap-location: bottom
    theme: cosmo
    grid:
      body-width: 1000px
      sidebar-width: 260px
      margin-width: 140px
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "results"
  prefix: ""
  config_json: ""
  cohort: ""
---

<!--
=============================================================================
genotype_qc_report.qmd

Renders a standalone HTML report from a genotype_qc_preimputation run that has
been organised by src/collect_cromwell_results.bash.

USAGE
  quarto render src/genotype_qc_report.qmd \
      -P results_dir=/path/to/results/genotype_qc_preimputation_YYYY-MM-DD \
      -P cohort="My cohort" \
      -P config_json=config/genotype_qc_preimputation_inputs.json \
      --output-dir /path/to/where/the/html/should/go

PARAMETERS
  results_dir  collected results directory (required in practice; relative
               paths are resolved from the .qmd's directory and then upwards)
  cohort       free-text cohort name shown in the header (optional)
  config_json  workflow inputs JSON, to document the thresholds (optional)
  prefix       output prefix; auto-detected from *_pipeline.log if omitted

Every file is located by pattern, not by name, so the report does not need to
know the run's prefix. Anything a run did not produce is reported as absent
instead of breaking the render — a partial results directory still renders.
=============================================================================
-->

```{r setup}
#| include: false

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(stringr)
  library(knitr)
})

knitr::opts_chunk$set(
  dev       = "png",
  dpi       = 150,
  fig.width = 8,
  fig.retina = 2
)

## ---------------------------------------------------------------------------
## Palette
##
## Two categorical slots only (blue / orange). Validated on a white surface for
## all pairs: CVD deltaE 24.7, normal-vision deltaE 33.6, both above 3:1
## contrast. Charts that would need more than two hues use small multiples or a
## single hue with the categories on an axis instead, so identity is never
## carried by colour alone. Grey is reserved for context/"other", never a series.
## ---------------------------------------------------------------------------
PAL <- c(blue = "#2a78d6", orange = "#eb6834", grey = "#8a8a86",
         ink = "#0b0b0b", ink2 = "#52514e", grid = "#e6e6e3")

theme_report <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 12, colour = PAL[["ink"]]),
      plot.subtitle    = element_text(size = 10, colour = PAL[["ink2"]]),
      plot.caption     = element_text(size = 8,  colour = PAL[["ink2"]], hjust = 0),
      axis.title       = element_text(size = 10, colour = PAL[["ink2"]]),
      axis.text        = element_text(size = 9,  colour = PAL[["ink2"]]),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = PAL[["grid"]], linewidth = 0.3),
      strip.text       = element_text(face = "bold", size = 10, colour = PAL[["ink"]]),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = 9, colour = PAL[["ink2"]])
    )
}
theme_set(theme_report())

## ---------------------------------------------------------------------------
## Locating result files
##
## Everything is discovered by pattern under the collected results directory
## (the layout written by src/collect_cromwell_results.bash), so the report does
## not need to know the run's output prefix. Any file a run did not produce is
## reported as absent rather than breaking the render.
## ---------------------------------------------------------------------------
## Quarto renders with the working directory set to the .qmd's own directory,
## so a relative path given on the command line is interpreted from src/.
## Try it as given, then walk up towards the project root.
## A "~" reaches R unexpanded when the shell did not expand it, so do it here.
resolve_path <- function(p) {
  if (!nzchar(p)) return(p)
  p <- path.expand(p)
  for (cand in c(p, file.path("..", p), file.path("..", "..", p))) {
    if (file.exists(cand) || dir.exists(cand)) return(normalizePath(cand, mustWork = FALSE))
  }
  normalizePath(p, mustWork = FALSE)
}

RD <- resolve_path(params$results_dir)
if (!dir.exists(RD)) {
  stop("results_dir not found: ", RD, "\n",
       "  Give an ABSOLUTE path — Quarto renders with the working directory set\n",
       "  to the .qmd's own directory, not the one you invoked it from:\n\n",
       "    quarto render /path/to/genotype_qc_pipeline/src/genotype_qc_report.qmd \\\n",
       "        -P results_dir=/absolute/path/to/results/genotype_qc_preimputation_YYYY-MM-DD \\\n",
       "        --output-dir /absolute/path/for/the/html\n")
}

picks <- function(subdir, pattern) {
  d <- if (nzchar(subdir)) file.path(RD, subdir) else RD
  if (!dir.exists(d)) return(character(0))
  sort(list.files(d, pattern = pattern, full.names = TRUE))
}
pick <- function(subdir, pattern) {
  h <- picks(subdir, pattern)
  if (length(h) == 0) NA_character_ else h[1]
}
have <- function(p) length(p) == 1 && !is.na(p) && file.exists(p)

rel  <- function(p) sub(paste0("^", RD, .Platform$file.sep), "", p, fixed = FALSE)
fref <- function(p) if (have(p)) paste0("`", rel(p), "`") else "_(not produced)_"

nlines <- function(p) if (have(p)) length(read_lines(p, progress = FALSE)) else NA_integer_
fsize  <- function(p) if (have(p)) file.info(p)$size else NA_real_

## Non-breaking space so "512 KB" never wraps across two table lines
hsize <- function(b) {
  ifelse(is.na(b), "—",
    ifelse(b >= 1024^3, sprintf("%.1f GB", b / 1024^3),
      ifelse(b >= 1024^2, sprintf("%.1f MB", b / 1024^2),
        ifelse(b >= 1024,   sprintf("%.0f KB", b / 1024), sprintf("%.0f B", b)))))
}

## ---------------------------------------------------------------------------
## Output helpers (used inside results="asis" chunks)
## ---------------------------------------------------------------------------
n_fmt <- function(x) ifelse(is.na(x), "—", formatC(as.numeric(x), format = "d", big.mark = ","))
p_fmt <- function(x) ifelse(is.na(x), "—", sprintf("%.1f%%", 100 * x))

show_table <- function(df, caption = NULL, align = NULL) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  ## Sample and family identifiers are numeric in PLINK files but are not
  ## quantities — keep them out of the thousands-separator formatting.
  id_cols <- c("FID", "IID", "FID1", "IID1", "FID2", "IID2", "ID", "PC")
  df <- df |> mutate(across(any_of(id_cols), as.character))
  cat(knitr::kable(df, caption = caption, align = align,
                   format.args = list(big.mark = ",")), sep = "\n")
  cat("\n\n")
}

## Plots produced by the pipeline live outside the report's directory, so they
## are inlined as base64 data URIs rather than linked — this keeps the rendered
## HTML a single self-contained file that can be emailed as-is.
show_image <- function(p, caption = "") {
  if (!have(p)) return(invisible(NULL))
  uri <- tryCatch(knitr::image_uri(p), error = function(e) p)
  cat(sprintf("![%s](%s)\n\n", caption, uri))
}

absent <- function(msg) {
  cat("\n::: {.callout-warning appearance='simple'}\n**Not available in this run.** ",
      msg, "\n:::\n\n", sep = "")
}

note <- function(msg) {
  cat("\n::: {.callout-note appearance='simple'}\n", msg, "\n:::\n\n", sep = "")
}

collapsible <- function(title, text) {
  cat("\n::: {.callout-note collapse='true'}\n## ", title, "\n\n```\n", sep = "")
  cat(text, sep = "\n")
  cat("\n```\n:::\n\n")
}

## ---------------------------------------------------------------------------
## The pipeline log — the backbone of the report
## ---------------------------------------------------------------------------
log_file <- pick("", "_pipeline\\.log$")
log_txt  <- if (have(log_file)) read_lines(log_file, progress = FALSE) else character(0)

prefix <- params$prefix
if (!nzchar(prefix) && have(log_file)) {
  prefix <- sub("_pipeline\\.log$", "", basename(log_file))
}

run_version <- str_match(paste(log_txt, collapse = "\n"),
                         "genotype_qc_preimputation\\s+v(\\S+)")[, 2]
run_date    <- str_match(paste(log_txt, collapse = "\n"),
                         "Run date\\s*:\\s*(\\S+ \\S+)")[, 2]

## Step lines carrying "SNPs: n  Samples: n"
step_counts <- tibble(line = log_txt) |>
  filter(str_detect(line, "SNPs:\\s*\\d+\\s+Samples:\\s*\\d+")) |>
  mutate(
    step    = str_trim(str_extract(line, "^.*?(?=\\s+SNPs:)")),
    snps    = as.integer(str_match(line, "SNPs:\\s*(\\d+)")[, 2]),
    samples = as.integer(str_match(line, "Samples:\\s*(\\d+)")[, 2])
  ) |>
  select(step, snps, samples) |>
  mutate(order = row_number())

first_step <- if (nrow(step_counts)) slice(step_counts, 1) else NULL
last_step  <- if (nrow(step_counts)) slice(step_counts, n()) else NULL

## Free-text log lines that are not SNP/sample counts (ancestry, LD, PCA, ...)
log_notes <- log_txt[str_detect(log_txt, "^Step ") &
                     !str_detect(log_txt, "SNPs:\\s*\\d+\\s+Samples:")]

## ---------------------------------------------------------------------------
## Common tables read once and reused
## ---------------------------------------------------------------------------
qc_file  <- pick("reports", "_sample_qc_status\\.tsv$")
qc_tbl   <- if (have(qc_file)) suppressWarnings(read_tsv(qc_file, show_col_types = FALSE,
                                                         progress = FALSE)) else NULL

anc_file <- pick("reports", "ancestry_assignments\\.tsv$")
if (!have(anc_file)) anc_file <- pick("pca", "ancestry_assignments\\.tsv$")
anc_tbl  <- if (have(anc_file)) suppressWarnings(read_tsv(anc_file, show_col_types = FALSE,
                                                          progress = FALSE)) else NULL

pc_cols <- function(df) if (is.null(df)) character(0) else grep("^PC\\d+$", names(df), value = TRUE)

## Sample lists come in three shapes: PLINK 2 `.id` files with a "#FID IID"
## header, the pipeline's own lists with a plain "FID IID" header, and bare
## two-column lists with no header at all.
read_id_file <- function(p) {
  if (!have(p)) return(NULL)
  x <- read_lines(p, progress = FALSE)
  x <- x[nzchar(str_trim(x))]
  if (length(x) == 0) return(tibble(FID = character(), IID = character()))
  if (str_detect(x[1], "^#") || str_detect(x[1], "^\\s*#?FID\\b")) x <- x[-1]
  if (length(x) == 0) return(tibble(FID = character(), IID = character()))
  parts <- str_split(str_trim(x), "\\s+", simplify = TRUE)
  if (ncol(parts) == 1) tibble(FID = parts[, 1], IID = parts[, 1])
  else tibble(FID = parts[, 1], IID = parts[, 2])
}

## Variants-per-chromosome report -> tidy table.
## Lines look like "  Chromosome  1:   48604 variants" or
## "  Chromosome 23 (X):   26957 variants" — keep the parenthetical annotation.
read_varchr <- function(p) {
  if (!have(p)) return(NULL)
  x <- read_lines(p, progress = FALSE)
  m <- str_match(x, "Chromosome\\s+(\\d+)\\s*(\\([A-Za-z]+\\))?\\s*:\\s+(\\d+)\\s+variants")
  ok <- !is.na(m[, 1])
  if (!any(ok)) return(NULL)
  ## Short axis labels: "23 (X)" collides with its neighbours on a 26-bar axis.
  tibble(chr = as.integer(m[ok, 2]), tag = m[ok, 3],
         variants = as.integer(m[ok, 4])) |>
    arrange(chr) |>
    mutate(chr_label = ifelse(is.na(tag), as.character(chr),
                              str_remove_all(tag, "[()]"))) |>
    select(chr, chr_label, variants)
}
```

```{r headline}
#| results: asis

## Assemble only the fields this run actually has — with no pipeline log there
## is no prefix, version or run date to show.
bits <- character(0)
if (nzchar(params$cohort))  bits <- c(bits, paste0("**Cohort:** ", params$cohort))
if (nzchar(prefix))         bits <- c(bits, paste0("**Run:** `", prefix, "`"))
if (!is.na(run_version))    bits <- c(bits, paste0("**pipeline version** ", run_version))
if (!is.na(run_date))       bits <- c(bits, paste0("**run date** ", run_date))
bits <- c(bits, paste0("**results** `", RD, "`"))

cat("::: {.callout-tip appearance='simple'}\n")
cat(paste(bits, collapse = "  ·  "), "\n", sep = "")
cat(":::\n\n")
```

This report summarises one execution of the `genotype_qc_preimputation` WDL
workflow: what was filtered at each step, how many variants and samples
survived, what the diagnostic plots show, and where every deliverable file
sits in the collected results directory. Numbers are read directly from the
run's own outputs — nothing here is recomputed from the genotypes.

# Headline numbers

```{r headline-table}
#| results: asis

rows <- list()
add <- function(metric, value, source) rows[[length(rows) + 1]] <<-
  tibble(Metric = metric, Value = value, `Read from` = source)

if (!is.null(first_step)) {
  add("Variants entering QC", n_fmt(first_step$snps), rel(log_file))
  add("Samples entering QC",  n_fmt(first_step$samples), rel(log_file))
}
if (!is.null(last_step)) {
  add("Variants in the final deliverable", n_fmt(last_step$snps), rel(log_file))
  add("Samples in the final deliverable",  n_fmt(last_step$samples), rel(log_file))
}
if (!is.null(first_step) && !is.null(last_step)) {
  add("Variants retained", p_fmt(last_step$snps / first_step$snps), "derived")
  add("Samples retained",  p_fmt(last_step$samples / first_step$samples), "derived")
}

rel_out <- read_id_file(pick("reports", "relatedness_flagged_samples\\.txt$"))
if (!is.null(rel_out)) add("Samples flagged as related (KING)", n_fmt(nrow(rel_out)),
                           "reports/relatedness_flagged_samples.txt")

het_fail <- pick("reports", "het_fail_samples\\.txt$")
if (have(het_fail)) add("Heterozygosity outliers removed",
                        n_fmt(nrow(read_id_file(het_fail))),
                        "reports/het_fail_samples.txt")

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  top <- anc_tbl |> count(superpop, sort = TRUE)
  add("Ancestry composition",
      paste(sprintf("%s: %s", top$superpop, n_fmt(top$n)), collapse = ", "),
      basename(anc_file))
}

vcfs <- picks("vcfs", "\\.vcf\\.gz$")
if (length(vcfs)) add("Imputation-ready VCFs", sprintf("%d files (%s)",
                       length(vcfs), hsize(sum(file.info(vcfs)$size))), "vcfs/")

if (length(rows)) {
  show_table(bind_rows(rows), align = c("l", "r", "l"))
} else {
  absent("No pipeline log was found, so headline numbers could not be derived.")
}
```

# Filtering trajectory

The workflow chains PLINK filters, writing a new `bed/bim/fam` triplet at each
step. The table below is the per-step count log the run wrote itself; the chart
shows the same numbers as a retention trajectory. Sharp drops are where QC did
its work — the interesting question at each step is whether the drop matches the
threshold that was applied.

```{r trajectory}
#| results: asis
#| fig-height: 6.5

if (nrow(step_counts) == 0) {
  absent(paste("No `*_pipeline.log` in", RD, "— the whole trajectory section",
               "depends on it. Check that call-WriteLog completed."))
} else {
  show_table(
    step_counts |>
      mutate(
        `Δ variants` = c(NA, diff(snps)),
        `Δ samples`  = c(NA, diff(samples))
      ) |>
      select(Step = step, Variants = snps, `Δ variants`,
             Samples = samples, `Δ samples`),
    caption = "Per-step variant and sample counts (source: the run's pipeline log)."
  )
}
```

```{r trajectory-plot}
#| fig-height: 6.5

if (nrow(step_counts) > 1) {
  traj <- step_counts |>
    pivot_longer(c(snps, samples), names_to = "metric", values_to = "n") |>
    mutate(
      metric = factor(recode(metric, snps = "Variants", samples = "Samples"),
                      levels = c("Variants", "Samples")),
      label  = str_trunc(str_squish(str_remove(step, "^Step\\s+\\S+\\s*")), 26),
      label  = factor(sprintf("%02d %s", order, label),
                      levels = sprintf("%02d %s", order, label)[!duplicated(order)])
    )

  ggplot(traj, aes(x = label, y = n, group = metric)) +
    geom_line(colour = PAL[["blue"]], linewidth = 0.7) +
    geom_point(colour = PAL[["blue"]], size = 2.2) +
    geom_text(aes(label = formatC(n, format = "d", big.mark = ",")),
              vjust = -0.9, size = 2.7, colour = PAL[["ink2"]]) +
    facet_wrap(~metric, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0.08, 0.18))) +
    labs(title = "Variants and samples retained through QC",
         subtitle = "Each panel has its own scale — the two measures are not comparable in size",
         x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
}
```

```{r trajectory-notes}
#| results: asis

if (length(log_notes)) {
  cat("Steps that report something other than a count — ancestry assignment,",
      "LD pruning, PCA, per-chromosome totals — logged the following:\n\n")
  collapsible("Additional log lines", log_notes)
}

if (length(log_txt)) collapsible("Full pipeline log", log_txt)
```

# Missingness filters (steps 1–2)

Step 1 (`--geno`) drops variants genotyped in too few samples; step 2
(`--mind`) drops samples genotyped at too few variants. Order matters: variants
are cleaned first so that a handful of failed assays cannot condemn otherwise
good samples.

The threshold sweep below is informational — the pipeline does not choose
thresholds from it. It shows how many variants and samples each candidate
threshold would have cost, so you can confirm the configured values sit on the
flat part of the curve rather than on a cliff.

```{r threshold-sweep}
#| results: asis

p <- pick("plots", "threshold_sweep\\.png$")
if (have(p)) {
  show_image(p, "Variant and sample retention across candidate --geno / --mind thresholds")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No `plots/threshold_sweep.png` — the `ThresholdSweep` task did not run or produced no plot.")
}
```

## Variants per chromosome, before imputation prep

```{r varchr-initial}
#| results: asis
#| fig-height: 3.6

vc_init <- read_varchr(pick("reports", "initial_variants_per_chr\\.txt$"))
if (is.null(vc_init)) {
  absent("No `reports/initial_variants_per_chr.txt`.")
} else {
  cat("Counted on the post-missingness dataset. Source file: ",
      fref(pick("reports", "initial_variants_per_chr\\.txt$")), "\n\n", sep = "")
}
```

```{r varchr-initial-plot}
#| fig-height: 3.6

if (!is.null(vc_init)) {
  ggplot(vc_init, aes(x = factor(chr_label, levels = chr_label), y = variants)) +
    geom_col(fill = PAL[["blue"]], width = 0.72) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
    labs(title = "Variants per chromosome entering QC",
         x = "Chromosome", y = "Variants")
}
```

```{r varchr-initial-table}
#| results: asis

if (!is.null(vc_init)) {
  show_table(
    vc_init |> select(Chromosome = chr_label, Variants = variants) |>
      bind_rows(tibble(Chromosome = "Total", Variants = sum(vc_init$variants))),
    caption = "Variants per chromosome before imputation preparation."
  )
}
```

# Sex check (step 3)

X-chromosome heterozygosity is compared with the sex recorded in the `.fam`
file. Samples whose F-statistic contradicts the recorded sex are removed — a
mismatch usually means a sample swap, which invalidates every phenotype
attached to it. The step is skipped automatically when the array carries no X
markers.

```{r sexcheck}
#| results: asis

## Two .sexcheck files can be present: PLINK's raw output and the pipeline's
## annotated copy. Prefer the annotated one.
sx <- pick("reports", "_annotated\\.sexcheck$")
if (!have(sx)) sx <- pick("reports", "\\.sexcheck$")
sx_fail <- pick("reports", "sex_fail_samples\\.fam$")

if (!have(sx)) {
  skipped <- any(str_detect(log_txt, regex("sex check.*skipped", ignore_case = TRUE)))
  if (skipped) {
    note(paste("**Sex check skipped.** The dataset contained no X-chromosome",
               "SNPs, so no F-statistic could be computed. This is expected for",
               "autosome-only arrays and does not indicate a problem."))
  } else {
    absent("No `reports/*.sexcheck` file was collected.")
  }
} else {
  ## The annotated report starts with "##" provenance lines (chromosome coding,
  ## number of X SNPs used) before the PLINK header.
  hdr <- read_lines(sx, n_max = 10, progress = FALSE)
  hdr <- str_trim(str_remove(hdr[str_starts(hdr, "##")], "^##\\s*"))
  sxt <- suppressWarnings(read_table(sx, comment = "##", show_col_types = FALSE,
                                     progress = FALSE))
  cat("Source file: ", fref(sx), "\n\n", sep = "")
  if (length(hdr)) cat(paste0("- ", hdr, collapse = "\n"), "\n\n", sep = "")

  if ("STATUS" %in% names(sxt)) {
    show_table(sxt |> count(STATUS, name = "Samples"),
               caption = "PLINK --check-sex status counts.")
    prob <- sxt |> filter(!str_detect(STATUS, regex("^ok$", ignore_case = TRUE)))
    if (nrow(prob)) {
      cat("**", n_fmt(nrow(prob)), " sample(s) were discordant and removed.**\n\n", sep = "")
      show_table(head(prob, 50))
      if (nrow(prob) > 50) cat("_(showing the first 50 of ", nrow(prob), ")_\n\n", sep = "")
    } else {
      cat("No sex discordances were found.\n\n")
    }
  } else {
    show_table(head(sxt, 20), caption = "Head of the sex check report.")
  }

  if (have(sx_fail)) {
    note(paste0("`", rel(sx_fail), "` is the `.fam` of the dataset **after** the ",
                "discordant samples were removed (", n_fmt(nlines(sx_fail)),
                " samples retained), not a list of failures — the name given to ",
                "it by `collect_cromwell_results.bash` is misleading."))
  }
}
```

# Ancestry (step 4)

Principal components are computed on the **1000 Genomes reference panel alone**,
and the study samples are then projected onto those axes with PLINK 2
`--score`. Because no study sample enters the eigendecomposition, related or
over-represented individuals cannot bend the axes — the coordinate system is
fixed by the reference and identical between runs. A random forest trained on
the labelled reference samples assigns each study sample a superpopulation and
a probability; samples below the configured probability threshold are labelled
`unassigned` rather than forced into a cluster.

These labels drive the ancestry-aware filtering that follows: HWE is tested
within ancestry, heterozygosity is judged within ancestry, and the step 13
subset is defined by them.

```{r ancestry-plot}
#| results: asis

p <- pick("plots", "_ancestry_pca\\.png$")
if (have(p)) {
  show_image(p, "Study samples projected onto 1000 Genomes principal components")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No ancestry PCA plot in `plots/`.")
}
```

```{r ancestry-counts}
#| results: asis
#| fig-height: 3.2

if (is.null(anc_tbl)) {
  absent("No `*_ancestry_assignments.tsv` — ancestry summaries are unavailable.")
} else {
  cat("Source file: ", fref(anc_file), " (columns: ",
      paste0("`", names(anc_tbl), "`", collapse = ", "), ")\n\n", sep = "")
}
```

```{r ancestry-counts-plot}
#| fig-height: 3.2

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  ac <- anc_tbl |> count(superpop, name = "n") |> arrange(desc(n))
  ggplot(ac, aes(x = reorder(superpop, n), y = n)) +
    geom_col(fill = PAL[["blue"]], width = 0.7) +
    geom_text(aes(label = scales::comma(n)), hjust = -0.2,
              size = 3, colour = PAL[["ink2"]]) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Assigned superpopulation",
         subtitle = "Counts are labelled directly; ancestry is on the axis, not in the colour",
         x = NULL, y = "Samples")
}
```

```{r ancestry-tables}
#| results: asis
#| fig-height: 3

if (!is.null(anc_tbl) && "superpop" %in% names(anc_tbl)) {
  ac <- anc_tbl |>
    count(superpop, name = "Samples") |>
    mutate(`% of cohort` = p_fmt(Samples / sum(Samples))) |>
    arrange(desc(Samples)) |>
    rename(Superpopulation = superpop)
  show_table(ac, caption = "Superpopulation assignment counts.")

  if ("probability" %in% names(anc_tbl)) {
    un <- sum(anc_tbl$superpop == "unassigned")
    if (un > 0) {
      note(sprintf(paste("%s sample(s) fell below the assignment probability",
                         "threshold and are labelled `unassigned`. They stay in the",
                         "dataset — they are simply excluded from ancestry-specific",
                         "tests and from the step 13 subset."), n_fmt(un)))
    }
  }
}
```

```{r ancestry-prob-plot}
#| fig-height: 3

if (!is.null(anc_tbl) && all(c("probability", "superpop") %in% names(anc_tbl))) {
  ggplot(anc_tbl, aes(x = probability)) +
    geom_histogram(bins = 40, fill = PAL[["blue"]], colour = "white", linewidth = 0.2) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
    labs(title = "Random forest assignment confidence",
         subtitle = "A mass near 1.0 means most samples sit cleanly inside a reference cluster",
         x = "Assignment probability", y = "Samples")
}
```

```{r ancestry-scree}
#| results: asis
#| fig-height: 3

ev_all <- picks("pca", "\\.eigenval$")
ev <- ev_all[!grepl("covariate", basename(ev_all))]
ev <- if (length(ev)) ev[1] else NA_character_
if (have(ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    scree <- tibble(PC = seq_along(vals), eigenvalue = vals,
                    pct = vals / sum(vals))
    cat("Variance carried by each reference PC (source: ", fref(ev), "):\n\n", sep = "")
    show_table(
      scree |> mutate(`Variance explained` = p_fmt(pct)) |>
        select(PC, Eigenvalue = eigenvalue, `Variance explained`) |>
        mutate(Eigenvalue = round(Eigenvalue, 3)),
      caption = "Ancestry PCA eigenvalues."
    )
  }
}
```

# Minor allele count filter (step 5)

Rare variants are removed on the full cohort, since rarity is a property of the
whole sample, not of one ancestry. Monomorphic sites go first, then everything
below the configured MAC threshold. Low-count variants carry almost no
association information and are the main source of spurious HWE and imputation
failures.

```{r mac}
#| results: asis

p <- pick("plots", "mac_plot\\.png$")
if (have(p)) {
  show_image(p, "Minor allele count distribution before and after filtering")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No `plots/mac_plot.png`.")
}

fb <- pick("reports", "^freq_before\\.frq$")
fa <- pick("reports", "^freq_after\\.frq$")
if (have(fb) || have(fa)) {
  ## These files carry one row per variant (tens of MB in a real run), so only
  ## the MAF column is parsed.
  summarise_frq <- function(p, label) {
    if (!have(p)) return(NULL)
    d <- suppressWarnings(read_table(p, col_types = cols_only(MAF = col_double()),
                                     progress = FALSE))
    if (!"MAF" %in% names(d)) return(NULL)
    tibble(
      Stage          = label,
      Variants       = nrow(d),
      `Monomorphic`  = sum(d$MAF == 0, na.rm = TRUE),
      `MAF < 1%`     = sum(d$MAF < 0.01, na.rm = TRUE),
      `MAF < 5%`     = sum(d$MAF < 0.05, na.rm = TRUE),
      `Median MAF`   = round(median(d$MAF, na.rm = TRUE), 4)
    )
  }
  frq <- bind_rows(summarise_frq(fb, "Before MAC filter"),
                   summarise_frq(fa, "After MAC filter"))
  if (!is.null(frq) && nrow(frq)) {
    show_table(frq, caption = paste0("Allele frequency summary (sources: ",
                                     rel(fb), ", ", rel(fa), ")."))
  }
} else {
  absent("No `reports/freq_before.frq` / `freq_after.frq`.")
}
```

# Hardy–Weinberg equilibrium (step 6)

HWE violations are **detected within each tested ancestry** and then removed
from the full cohort. Testing the pooled cohort would flag variants that are
perfectly in equilibrium within each population and out of it only because the
cohort mixes populations — the Wahlund effect. Detecting per ancestry and
removing globally keeps one variant set across the whole dataset while avoiding
that artefact.

```{r hwe}
#| results: asis

hwe_line <- log_txt[str_detect(log_txt, regex("^Step 6\\s+HWE", ignore_case = TRUE))]
mac_line <- log_txt[str_detect(log_txt, regex("^Step 5", ignore_case = TRUE))]

if (length(hwe_line) && length(mac_line)) {
  n_before <- as.integer(str_match(mac_line[1], "SNPs:\\s*(\\d+)")[, 2])
  n_after  <- as.integer(str_match(hwe_line[1], "SNPs:\\s*(\\d+)")[, 2])
  if (!is.na(n_before) && !is.na(n_after)) {
    show_table(
      tibble(Metric = c("Variants before HWE filter", "Variants after HWE filter",
                        "Removed for HWE violation"),
             Value = n_fmt(c(n_before, n_after, n_before - n_after))),
      caption = "HWE filtering, from the pipeline log.", align = c("l", "r")
    )
  }
} else {
  absent("The pipeline log has no step 5/6 count lines to compare.")
}

hwe_files <- picks("reports", "\\.hwe$")
if (length(hwe_files)) {
  cat("Per-variant HWE statistics: ", paste(sprintf("`%s`", rel(hwe_files)),
      collapse = ", "), "\n\n", sep = "")
} else {
  note(paste("No `.hwe` table is collected: the workflow applies `--hwe` as a",
             "filter rather than running `--hardy`, so PLINK writes no per-variant",
             "HWE report. The counts above are the authoritative record of what",
             "the step removed."))
}
```

# Heterozygosity (steps 7a–7b)

Step 7a LD-prunes the dataset with high-LD regions excluded. The pruned SNP
list is a **helper only** — it feeds the heterozygosity and relatedness checks
and is never applied to the delivered genotypes.

Step 7b flags samples more than 3 SD from the mean heterozygosity rate.
Excess heterozygosity suggests contamination or a mixed sample; a deficit
suggests inbreeding or poor DNA quality. Detection and removal both happen
within the ancestry subset, so samples from untested populations are not judged
against a mean that does not describe them.

```{r het}
#| results: asis

p <- pick("plots", "heterozygosity\\.png$")
if (have(p)) {
  show_image(p, "Heterozygosity rate distribution with the ±3 SD outlier bounds")
  cat("Source file: ", fref(p), "\n\n", sep = "")
} else {
  absent("No `plots/heterozygosity.png`.")
}

het <- pick("reports", "R_check\\.het$")
hf  <- pick("reports", "het_fail_samples\\.txt$")

if (have(het)) {
  h <- suppressWarnings(read_table(het, show_col_types = FALSE, progress = FALSE))
  nm <- names(h)
  if (all(c("N(NM)", "O(HOM)") %in% nm) || all(c("N.NM.", "O.HOM.") %in% nm)) {
    n_nm  <- if ("N(NM)" %in% nm) h[["N(NM)"]] else h[["N.NM."]]
    o_hom <- if ("O(HOM)" %in% nm) h[["O(HOM)"]] else h[["O.HOM."]]
    rate  <- (n_nm - o_hom) / n_nm
    show_table(
      tibble(Metric = c("Samples tested", "Mean heterozygosity rate",
                        "SD", "Lower bound (mean − 3 SD)", "Upper bound (mean + 3 SD)"),
             Value  = c(n_fmt(length(rate)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE)),
                        sprintf("%.4f", sd(rate, na.rm = TRUE)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE) - 3 * sd(rate, na.rm = TRUE)),
                        sprintf("%.4f", mean(rate, na.rm = TRUE) + 3 * sd(rate, na.rm = TRUE)))),
      caption = paste0("Heterozygosity summary (source: ", rel(het), ")."),
      align = c("l", "r")
    )
  }
} else {
  absent("No `reports/R_check.het`.")
}

if (have(hf)) {
  n_fail <- nrow(read_id_file(hf))
  cat("**", n_fmt(n_fail), " sample(s) failed the heterozygosity check** and were removed. ",
      "Full list: ", fref(hf), "\n\n", sep = "")
  if (n_fail > 0 && n_fail <= 50) {
    ff <- read_id_file(hf)
    show_table(ff, caption = "Samples removed as heterozygosity outliers.")
  }
} else {
  absent("No `reports/het_fail_samples.txt`.")
}
```

# Chromosome filter (step 8)

The dataset is restricted to the configured chromosome range (`--chr 1-23` by
default). This is the **final QC dataset**: everything after it is
harmonisation and packaging, not filtering on QC grounds.

```{r chrfilter}
#| results: asis

l8 <- step_counts |> filter(str_detect(step, regex("^Step 8", ignore_case = TRUE)))
if (nrow(l8)) {
  show_table(l8 |> select(Step = step, Variants = snps, Samples = samples),
             caption = "Final QC dataset (step 8).")
  note(paste("The step 8 dataset stays in the Cromwell run directory and is",
             "**not** collected into `results/`. What lands in `qc_dataset/` is the",
             "step 10 harmonised dataset, so that the PLINK files and the",
             "imputation VCFs describe exactly the same variants."))
} else {
  absent("No step 8 line in the pipeline log.")
}
```

# Relatedness (step 9)

KING-robust kinship is computed on the final QC dataset, reusing the step 7a
pruned SNP list. Running it here rather than earlier is deliberate: the
ancestry PCA takes its axes from the reference panel alone, so relatives could
not have distorted it, and computing kinship on the post-MAC, post-HWE variant
set gives a cleaner estimate.

**Related samples are flagged, not removed.** Which member of a pair to drop
depends on the analysis — phenotype completeness, case status, genotyping
quality — so the pipeline leaves the decision to you and ships both lists.

```{r related}
#| results: asis
#| fig-height: 3.2

rout <- pick("reports", "relatedness_flagged_samples\\.txt$")
rin  <- pick("reports", "relatedness_unrelated_samples\\.txt$")
gen  <- pick("reports", "\\.genome$")

rows <- list()
if (have(rout)) rows[[length(rows) + 1]] <-
  tibble(Metric = "Samples flagged as related (to remove for an unrelated set)",
         Value = n_fmt(nrow(read_id_file(rout))), File = rel(rout))
if (have(rin)) rows[[length(rows) + 1]] <-
  tibble(Metric = "Samples in the maximal unrelated set",
         Value = n_fmt(nrow(read_id_file(rin))), File = rel(rin))

gtab <- NULL
if (have(gen)) {
  gtab <- suppressWarnings(read_table(gen, show_col_types = FALSE, progress = FALSE))
  rows[[length(rows) + 1]] <-
    tibble(Metric = "Sample pairs above the pi-hat reporting threshold",
           Value = n_fmt(nrow(gtab)), File = rel(gen))
}

if (length(rows)) {
  show_table(bind_rows(rows), align = c("l", "r", "l"))
} else {
  absent("No relatedness outputs in `reports/`.")
}
```

```{r related-plot}
#| fig-height: 3.2

if (!is.null(gtab) && "PI_HAT" %in% names(gtab) && nrow(gtab) > 0) {
  ggplot(gtab, aes(x = PI_HAT)) +
    geom_histogram(bins = 40, fill = PAL[["blue"]], colour = "white", linewidth = 0.2) +
    geom_vline(xintercept = 0.5,   colour = PAL[["orange"]], linewidth = 0.6, linetype = "dashed") +
    geom_vline(xintercept = 0.25,  colour = PAL[["orange"]], linewidth = 0.6, linetype = "dashed") +
    annotate("text", x = 0.5,  y = Inf, label = "1st degree (0.5)", vjust = 1.6,
             hjust = -0.05, size = 2.8, colour = PAL[["orange"]]) +
    annotate("text", x = 0.25, y = Inf, label = "2nd degree (0.25)", vjust = 3.2,
             hjust = -0.05, size = 2.8, colour = PAL[["orange"]]) +
    scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
    labs(title = "Pi-hat distribution of reported pairs",
         subtitle = "Only pairs above the configured pihat_min appear in the .genome file",
         x = "PI_HAT (proportion IBD)", y = "Pairs")
}
```

```{r related-pairs}
#| results: asis

if (!is.null(gtab) && nrow(gtab) > 0) {
  cols <- intersect(c("FID1", "IID1", "FID2", "IID2", "Z0", "Z1", "Z2", "PI_HAT"),
                    names(gtab))
  top <- gtab |> arrange(desc(.data[["PI_HAT"]])) |> head(25) |> select(all_of(cols))
  show_table(top, caption = "The 25 most closely related pairs (full table in the .genome file).")
  if (nrow(gtab) > 25) cat("_(", n_fmt(nrow(gtab)), " pairs in total)_\n\n", sep = "")
}
```

# Imputation preparation (step 10)

Will Rayner's `HRC-1000G-check-bim.pl` harmonises the QC dataset against the
imputation reference panel: it fixes strand flips, aligns ref/alt, and drops
variants the panel cannot accept — sites absent from the panel, palindromic
A/T and C/G SNPs that cannot be strand-resolved, allele mismatches, and
variants whose frequency differs from the reference by more than 0.2.

The step emits one VCF per chromosome for upload **and** merges the harmonised
per-chromosome filesets back into a single `bed/bim/fam`. That combined file is
the PLINK equivalent of the VCFs — same variants, same alleles — which is what
makes the two deliverables directly comparable.

```{r checkbim}
#| results: asis

cb <- pick("reports", "check-bim\\.log$")
if (have(cb)) {
  txt <- read_lines(cb, progress = FALSE)
  m <- str_match(txt, "^\\s*([A-Za-z][^/]*?)\\s+(\\d+)\\s*$")
  ok <- !is.na(m[, 1]) & nchar(str_trim(m[, 2])) > 3
  if (any(ok)) {
    show_table(
      tibble(Check = str_trim(m[ok, 2]), Variants = as.integer(m[ok, 3])),
      caption = paste0("Counts parsed from ", rel(cb), ".")
    )
  }
  collapsible("check-bim.log (full)", txt)
} else {
  absent("No `reports/check-bim.log`.")
}
```

## Variants per chromosome in the imputation-ready data

```{r varchr-final}
#| results: asis
#| fig-height: 3.8

vc_fin <- read_varchr(pick("reports", "final_variants_per_chr\\.txt$"))
if (is.null(vc_fin)) {
  absent("No `reports/final_variants_per_chr.txt`.")
} else {
  cat("These counts match the VCFs variant-for-variant. Source file: ",
      fref(pick("reports", "final_variants_per_chr\\.txt$")), "\n\n", sep = "")
}
```

```{r varchr-final-plot}
#| fig-height: 3.8

if (!is.null(vc_fin)) {
  if (!is.null(vc_init)) {
    cmp <- bind_rows(
      vc_init |> mutate(stage = "Entering QC"),
      vc_fin  |> mutate(stage = "Imputation-ready")
    ) |>
      mutate(stage = factor(stage, levels = c("Entering QC", "Imputation-ready")),
             chr_label = factor(chr_label, levels = unique(vc_init$chr_label)))

    ggplot(cmp, aes(x = chr_label, y = variants, fill = stage)) +
      geom_col(position = position_dodge(width = 0.78), width = 0.68) +
      scale_fill_manual(values = unname(PAL[c("blue", "orange")])) +
      scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
      labs(title = "Variants per chromosome, before QC and after imputation prep",
           subtitle = "The gap is what missingness, MAC, HWE and panel harmonisation removed",
           x = "Chromosome", y = "Variants")
  } else {
    ggplot(vc_fin, aes(x = factor(chr_label, levels = chr_label), y = variants)) +
      geom_col(fill = PAL[["blue"]], width = 0.72) +
      scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
      labs(title = "Variants per chromosome in the imputation-ready data",
           x = "Chromosome", y = "Variants")
  }
}
```

```{r varchr-final-table}
#| results: asis

if (!is.null(vc_fin)) {
  tab <- vc_fin |> select(Chromosome = chr_label, `Imputation-ready` = variants)
  if (!is.null(vc_init)) {
    tab <- vc_init |>
      select(Chromosome = chr_label, `Entering QC` = variants) |>
      full_join(tab, by = "Chromosome") |>
      mutate(Removed = `Entering QC` - `Imputation-ready`,
             `Retained` = p_fmt(`Imputation-ready` / `Entering QC`))
  }
  show_table(tab, caption = "Per-chromosome variant counts.")
}
```

## Imputation VCFs

```{r vcfs}
#| results: asis

vcf_files <- picks("vcfs", "\\.vcf\\.gz$")
tbi_files <- picks("vcfs", "\\.vcf\\.gz\\.tbi$")

if (length(vcf_files) == 0) {
  absent("No VCFs in `vcfs/` — the imputation deliverable is missing.")
} else {
  vi <- tibble(File = basename(vcf_files), Bytes = file.info(vcf_files)$size) |>
    mutate(chr = as.integer(str_match(File, "chr(\\d+)\\.vcf\\.gz$")[, 2]),
           Size = hsize(Bytes),
           Index = ifelse(paste0(File, ".tbi") %in% basename(tbi_files), "yes", "MISSING")) |>
    arrange(chr) |>
    select(File, Size, `Tabix index` = Index)
  show_table(vi, caption = sprintf(
    "Imputation-ready VCFs in `vcfs/` (%d files, %s total).",
    length(vcf_files), hsize(sum(file.info(vcf_files)$size))))

  if (any(vi$`Tabix index` == "MISSING")) {
    absent("One or more VCFs have no `.tbi` index — imputation servers will reject them.")
  } else {
    note(paste("Upload the contents of `vcfs/` to the imputation server",
               "(HRC or TOPMed, matching the panel the run was harmonised against)."))
  }
}
```

# Combined imputation-ready dataset (step 11)

The same harmonised data in PLINK form, in both PLINK 1 and PLINK 2 encodings.
Use these for any analysis you want to run on the pre-imputation genotypes;
they are variant-for-variant identical to the uploaded VCFs.

```{r combined}
#| results: asis

fam <- pick("qc_dataset", "\\.fam$")
bim <- pick("qc_dataset", "\\.bim$")
bed <- pick("qc_dataset", "\\.bed$")
pgen <- pick("qc_dataset", "\\.pgen$")
pvar <- pick("qc_dataset", "\\.pvar$")
psam <- pick("qc_dataset", "\\.psam$")

if (!have(bim) && !have(pgen)) {
  absent("Nothing in `qc_dataset/` — the combined deliverable is missing.")
} else {
  rows <- list()
  if (have(bim)) rows[[length(rows) + 1]] <-
    tibble(Format = "PLINK 1 (bed/bim/fam)",
           Variants = n_fmt(nlines(bim)), Samples = n_fmt(nlines(fam)),
           Files = paste(sprintf("`%s`", basename(c(bed, bim, fam))), collapse = ", "))
  if (have(psam)) rows[[length(rows) + 1]] <-
    tibble(Format = "PLINK 2 (pgen/pvar/psam)",
           Variants = if (have(pvar)) n_fmt(sum(!str_starts(
                        read_lines(pvar, progress = FALSE), "#"))) else "—",
           Samples = n_fmt(max(0, nlines(psam) - 1)),
           Files = paste(sprintf("`%s`", basename(c(pgen, pvar, psam))), collapse = ", "))
  show_table(bind_rows(rows), caption = "Contents of `qc_dataset/`.")
}
```

# Covariate PCs (step 12)

Within-cohort principal components for use as association covariates. The axes
are estimated on the KING-unrelated samples — relatives would otherwise pull
components towards their own family structure — and then **all** samples are
projected onto them, so nobody is dropped from the covariate file.

These are a different object from the step 4 ancestry PCs: those place samples
against an external reference to label ancestry, these describe residual
structure inside this cohort.

```{r covar}
#| results: asis
#| fig-height: 3

cov_f  <- pick("pca", "_covariate_pca_covariates\\.tsv$")
if (!have(cov_f)) cov_f <- pick("pca", "_covariates\\.tsv$")
cov_ev <- pick("pca", "_covariate_pca\\.eigenval$")

cov_tbl <- if (have(cov_f)) suppressWarnings(read_tsv(cov_f, show_col_types = FALSE,
                                                      progress = FALSE)) else NULL

if (is.null(cov_tbl)) {
  absent("No covariate PC file in `pca/`.")
} else {
  npc <- length(pc_cols(cov_tbl))
  cat("**", n_fmt(nrow(cov_tbl)), " samples × ", npc, " PCs** — ", fref(cov_f),
      ". Columns: `FID`, `IID`, `PC1`…`PC", npc, "`.\n\n", sep = "")
}
```

```{r covar-scree}
#| fig-height: 3

scree <- NULL
if (have(cov_ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(cov_ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    scree <- tibble(PC = seq_along(vals), eigenvalue = vals, pct = vals / sum(vals))
    ggplot(scree, aes(x = PC, y = pct)) +
      geom_col(fill = PAL[["blue"]], width = 0.7) +
      geom_text(data = head(scree, 5), aes(label = p_fmt(pct)),
                vjust = -0.6, size = 2.8, colour = PAL[["ink2"]]) +
      scale_x_continuous(breaks = scales::pretty_breaks()) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
      labs(title = "Variance explained by each within-cohort PC",
           subtitle = "Where the bars flatten is a reasonable place to stop taking covariates",
           x = "Principal component", y = "Share of variance")
  }
}
```

```{r covar-scree-table}
#| results: asis

if (!is.null(scree)) {
  show_table(
    scree |> head(10) |>
      mutate(Eigenvalue = round(eigenvalue, 3), `Variance explained` = p_fmt(pct)) |>
      select(PC, Eigenvalue, `Variance explained`),
    caption = paste0("First 10 within-cohort eigenvalues (source: ", rel(cov_ev), ").")
  )
}
```

```{r covar-scatter}
#| fig-height: 5.5

## PC1/PC2 as small multiples by ancestry: identity comes from the facet strip,
## so the chart needs no categorical palette and stays readable under CVD.
if (!is.null(qc_tbl) && all(c("PC1", "PC2", "ancestry") %in% names(qc_tbl))) {
  d <- qc_tbl |> filter(!is.na(PC1), !is.na(PC2))
  if (nrow(d) > 0 && n_distinct(d$ancestry) > 1) {
    ggplot(d, aes(x = PC1, y = PC2)) +
      geom_point(data = transform(d, ancestry = NULL), colour = PAL[["grey"]],
                 alpha = 0.35, size = 0.5) +
      geom_point(colour = PAL[["blue"]], alpha = 0.75, size = 0.7) +
      facet_wrap(~ancestry) +
      labs(title = "Within-cohort PC1 vs PC2, by assigned ancestry",
           subtitle = "Grey shows the whole cohort in every panel for context",
           x = "PC1", y = "PC2")
  } else {
    ggplot(d, aes(x = PC1, y = PC2)) +
      geom_point(colour = PAL[["blue"]], alpha = 0.7, size = 0.8) +
      labs(title = "Within-cohort PC1 vs PC2", x = "PC1", y = "PC2")
  }
}
```

# Sample QC status table

One row per participant, joining ancestry, the relatedness flag and the
covariate PCs. This single file is enough to subset the cohort and pull
covariates without touching the genotype files — for example, keep
`ancestry == "EUR" & !related`, then pass `PC1..PC10` to your association model.

```{r qcstatus}
#| results: asis

if (is.null(qc_tbl)) {
  absent("No `reports/*_sample_qc_status.tsv`.")
} else {
  cat("Source file: ", fref(qc_file), " — ", n_fmt(nrow(qc_tbl)), " samples, ",
      "columns: ", paste0("`", names(qc_tbl)[seq_len(min(6, ncol(qc_tbl)))], "`",
      collapse = ", "),
      if (ncol(qc_tbl) > 6) sprintf(", … (%d columns in total)", ncol(qc_tbl)) else "",
      "\n\n", sep = "")

  ## The table is built from the step 4 ancestry assignments, which were made
  ## before the heterozygosity filter. If samples were dropped after step 4 the
  ## table lists more people than the delivered genotypes contain.
  n_final <- if (have(fam)) nlines(fam) else NA_integer_
  if (!is.na(n_final) && nrow(qc_tbl) != n_final) {
    note(sprintf(paste("This table has %s rows but the delivered dataset holds %s",
                       "samples. The extra rows are samples assigned an ancestry at",
                       "step 4 and dropped later (heterozygosity outliers, and sex",
                       "discordances if the check ran). Join on `FID`/`IID` against",
                       "the `.fam`/`.psam` before using it as a cohort list."),
                 n_fmt(nrow(qc_tbl)), n_fmt(n_final)))
  }

  if (all(c("ancestry", "related") %in% names(qc_tbl))) {
    ct <- qc_tbl |>
      mutate(related = ifelse(as.logical(related), "Related (flagged)", "Unrelated")) |>
      count(ancestry, related) |>
      pivot_wider(names_from = related, values_from = n, values_fill = 0)
    if (!"Related (flagged)" %in% names(ct)) ct[["Related (flagged)"]] <- 0L
    if (!"Unrelated" %in% names(ct)) ct[["Unrelated"]] <- 0L
    ct <- ct |>
      mutate(Total = Unrelated + `Related (flagged)`) |>
      arrange(desc(Total)) |>
      rename(Ancestry = ancestry)
    ct <- bind_rows(ct, summarise(ct, Ancestry = "All",
                                  across(where(is.numeric), sum)))
    show_table(ct, caption = "Cohort composition by ancestry and relatedness flag.")
  }

  ## Show the leading columns only — with 20+ PCs the full table does not fit
  ## on a page, and the point here is the shape of the file, not its contents.
  head_cols <- intersect(c("FID", "IID", "ancestry", "ancestry_prob", "related",
                           "PC1", "PC2", "PC3"), names(qc_tbl))
  show_table(head(qc_tbl, 5) |> select(all_of(head_cols)) |>
               mutate(across(where(is.numeric), \(x) round(x, 4))),
             caption = paste0("First five rows of the sample QC status table",
                              if (ncol(qc_tbl) > length(head_cols))
                                sprintf(" (first %d of %d columns).",
                                        length(head_cols), ncol(qc_tbl)) else "."))
}
```

# Unrelated single-ancestry subset (step 13)

An additional deliverable: the imputation-ready data restricted to samples of
one superpopulation that are not KING-flagged as related, with covariate PCs
recomputed **inside** the subset. Whole-cohort PCs largely capture
between-ancestry structure, which is not the axis you want to adjust for in a
within-population analysis — hence the separate PCA.

```{r subset}
#| results: asis
#| fig-height: 3

s_bim  <- pick("subset", "\\.bim$")
s_fam  <- pick("subset", "\\.fam$")
s_bed  <- pick("subset", "\\.bed$")
s_pgen <- pick("subset", "\\.pgen$")
s_pvar <- pick("subset", "\\.pvar$")
s_psam <- pick("subset", "\\.psam$")
s_keep <- pick("subset", "_keep\\.txt$")
s_cov  <- pick("subset", "_covariates\\.tsv$")
s_ev   <- pick("subset", "\\.eigenval$")

subset_pop <- if (have(s_keep)) str_match(basename(s_keep),
                  "_unrelated_([A-Za-z]+)_keep\\.txt$")[, 2] else NA_character_

if (!have(s_bim) && !have(s_keep)) {
  absent("Nothing in `subset/` — step 13 did not run or was not collected.")
} else {
  if (!is.na(subset_pop)) {
    cat("Subset population: **", subset_pop, "**, unrelated samples only.\n\n", sep = "")
  }
  rows <- list()
  if (have(s_keep)) rows[[length(rows) + 1]] <-
    tibble(Item = "Samples in the subset", Value = n_fmt(nlines(s_keep)), File = rel(s_keep))
  if (have(s_bim)) rows[[length(rows) + 1]] <-
    tibble(Item = "Variants (PLINK 1)", Value = n_fmt(nlines(s_bim)),
           File = paste(basename(c(s_bed, s_bim, s_fam)), collapse = ", "))
  if (have(s_psam)) rows[[length(rows) + 1]] <-
    tibble(Item = "PLINK 2 copy", Value = "pgen/pvar/psam",
           File = paste(basename(c(s_pgen, s_pvar, s_psam)), collapse = ", "))
  if (have(s_cov)) {
    sc <- suppressWarnings(read_tsv(s_cov, show_col_types = FALSE, progress = FALSE))
    rows[[length(rows) + 1]] <-
      tibble(Item = "Within-subset covariate PCs",
             Value = sprintf("%s samples × %d PCs", n_fmt(nrow(sc)), length(pc_cols(sc))),
             File = rel(s_cov))
  }
  show_table(bind_rows(rows), caption = "Contents of `subset/`.", align = c("l", "r", "l"))
}
```

```{r subset-scree}
#| fig-height: 3

if (have(s_ev)) {
  vals <- suppressWarnings(as.numeric(read_lines(s_ev, progress = FALSE)))
  vals <- vals[!is.na(vals)]
  if (length(vals)) {
    sc <- tibble(PC = seq_along(vals), pct = vals / sum(vals))
    ggplot(sc, aes(x = PC, y = pct)) +
      geom_col(fill = PAL[["blue"]], width = 0.7) +
      geom_text(data = head(sc, 5), aes(label = p_fmt(pct)),
                vjust = -0.6, size = 2.8, colour = PAL[["ink2"]]) +
      scale_x_continuous(breaks = scales::pretty_breaks()) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
      labs(title = paste0("Variance explained within the ",
                          ifelse(is.na(subset_pop), "", paste0(subset_pop, " ")),
                          "unrelated subset"),
           subtitle = "Flatter than the whole-cohort scree — between-ancestry structure is gone",
           x = "Principal component", y = "Share of variance")
  }
}
```

# What to use for what

```{r deliverables}
#| results: asis

d <- tribble(
  ~Purpose, ~Location,
  "Upload to the imputation server (HRC / TOPMed)", "`vcfs/*.vcf.gz` + `.tbi`",
  "Pre-imputation analysis in PLINK 1", "`qc_dataset/*_combined.{bed,bim,fam}`",
  "Pre-imputation analysis in PLINK 2", "`qc_dataset/*.{pgen,pvar,psam}`",
  "Association covariates for the whole cohort", "`pca/*_covariate_pca_covariates.tsv`",
  "Subsetting the cohort by ancestry or relatedness", "`reports/*_sample_qc_status.tsv`",
  "Dropping relatives from an analysis", "`reports/relatedness_flagged_samples.txt` with PLINK `--remove`",
  "A ready-made unrelated single-ancestry analysis set", "`subset/`",
  "Per-step counts for a methods section", "the `*_pipeline.log` at the top level"
)
show_table(d)

note(paste("Related samples are **flagged, not removed**, everywhere except the",
           "step 13 subset. If your analysis assumes unrelated individuals, apply",
           "`--remove reports/relatedness_flagged_samples.txt` yourself."))
```

# File inventory

```{r inventory}
#| results: asis

all_files <- list.files(RD, recursive = TRUE, full.names = TRUE)
if (length(all_files) == 0) {
  absent("The results directory is empty.")
} else {
  describe <- function(f) {
    b <- basename(f)
    case_when(
      str_detect(b, "_pipeline\\.log$")                ~ "Per-step SNP/sample counts for the whole run",
      str_detect(b, "\\.vcf\\.gz$")                    ~ "Imputation-ready VCF (one per chromosome)",
      str_detect(b, "\\.vcf\\.gz\\.tbi$")              ~ "Tabix index for the VCF",
      str_detect(b, "_combined\\.(bed|bim|fam)$")      ~ "Combined harmonised dataset, PLINK 1",
      str_detect(b, "\\.(pgen|pvar|psam)$") & str_detect(f, "/subset/") ~ "Unrelated subset, PLINK 2",
      str_detect(b, "\\.(bed|bim|fam)$")    & str_detect(f, "/subset/") ~ "Unrelated subset, PLINK 1",
      str_detect(b, "\\.(pgen|pvar|psam)$")            ~ "Combined harmonised dataset, PLINK 2",
      str_detect(b, "_keep\\.txt$")                    ~ "FID/IID of the unrelated single-ancestry subset",
      str_detect(b, "_sample_qc_status\\.tsv$")        ~ "Per-sample ancestry, relatedness flag and PCs",
      str_detect(b, "ancestry_assignments\\.tsv$")     ~ "Assigned superpopulation and probability",
      str_detect(b, "_ancestry_pca\\.png$")            ~ "Study samples projected on 1000G PCs",
      str_detect(b, "threshold_sweep\\.png$")          ~ "Retention across candidate thresholds",
      str_detect(b, "mac_plot\\.png$")                 ~ "MAC distribution before/after filtering",
      str_detect(b, "heterozygosity\\.png$")           ~ "Heterozygosity distribution and outlier bounds",
      str_detect(b, "relatedness_flagged_samples")     ~ "KING-flagged related samples (remove for unrelated set)",
      str_detect(b, "relatedness_unrelated_samples")   ~ "Maximal unrelated sample set",
      str_detect(b, "\\.genome$")                      ~ "Pairwise IBD estimates above pihat_min",
      str_detect(b, "R_check\\.het$")                  ~ "Per-sample heterozygosity statistics",
      str_detect(b, "het_fail_samples\\.txt$")         ~ "Heterozygosity outliers that were removed",
      str_detect(b, "_annotated\\.sexcheck$")          ~ "X-chromosome sex check report (annotated)",
      str_detect(b, "\\.sexcheck$")                    ~ "X-chromosome sex check report (raw PLINK output)",
      str_detect(b, "sex_fail_samples\\.fam$")         ~ "FAM of the dataset AFTER sex-discordant samples were removed",
      str_detect(b, "freq_before\\.frq$")              ~ "Allele frequencies before the MAC filter",
      str_detect(b, "freq_after\\.frq$")               ~ "Allele frequencies after the MAC filter",
      str_detect(b, "check-bim\\.log$")                ~ "HRC/TOPMed harmonisation log",
      str_detect(b, "initial_variants_per_chr")        ~ "Variants per chromosome entering QC",
      str_detect(b, "final_variants_per_chr")          ~ "Variants per chromosome in the VCFs",
      str_detect(b, "_covariates\\.tsv$")              ~ "PC covariates (FID IID PC1..PCn)",
      str_detect(b, "\\.eigenval$")                    ~ "PCA eigenvalues (variance per PC)",
      str_detect(b, "\\.eigenvec\\.var$")              ~ "PCA variant weights (used for projection)",
      str_detect(b, "\\.eigenvec$")                    ~ "PCA sample scores",
      str_detect(b, "\\.sscore$")                      ~ "Study samples projected onto reference PCs",
      str_detect(b, "prune\\.in$")                     ~ "LD-pruned SNP list",
      str_detect(b, "\\.(bed|bim|fam)$")               ~ "PLINK 1 genotype fileset",
      TRUE                                             ~ ""
    )
  }

  inv <- tibble(path = all_files) |>
    mutate(Directory = ifelse(dirname(path) == RD, "(top level)", basename(dirname(path))),
           File = basename(path),
           Size = hsize(file.info(path)$size),
           What = describe(path)) |>
    arrange(match(Directory, c("(top level)", "qc_dataset", "vcfs", "subset",
                               "pca", "reports", "plots")), File) |>
    select(Directory, File, Size, What)

  ## Collapse the per-chromosome VCF rows into one line each
  vcf_rows <- inv |> filter(str_detect(File, "\\.vcf\\.gz$"))
  tbi_rows <- inv |> filter(str_detect(File, "\\.vcf\\.gz\\.tbi$"))
  inv <- inv |> filter(!str_detect(File, "\\.vcf\\.gz"))
  if (nrow(vcf_rows)) {
    inv <- bind_rows(inv, tibble(
      Directory = "vcfs",
      File = sprintf("%d × *.vcf.gz (+ %d × .tbi)", nrow(vcf_rows), nrow(tbi_rows)),
      Size = hsize(sum(file.info(c(picks("vcfs", "\\.vcf\\.gz$"),
                                   picks("vcfs", "\\.tbi$")))$size)),
      What = "Imputation-ready VCFs and their indexes"))
  }

  show_table(inv, caption = sprintf("Every file in `%s` (%d files, %s).",
                                    RD, length(all_files),
                                    hsize(sum(file.info(all_files)$size))))
}
```

# Run parameters and provenance

```{r params}
#| results: asis

cfg_path <- resolve_path(params$config_json)
if (nzchar(cfg_path) && file.exists(cfg_path)) {
  if (requireNamespace("jsonlite", quietly = TRUE)) {
    cfg <- jsonlite::fromJSON(cfg_path)
    ct <- tibble(Parameter = sub("^genotype_qc_preimputation\\.", "", names(cfg)),
                 Value = vapply(cfg, \(x) paste(as.character(x), collapse = ", "), ""))
    ## Show the QC knobs up front; file and tool paths go in the fold.
    qc_params <- c("output_prefix", "geno_threshold", "mind_threshold",
                   "mac_threshold", "hwe_pvalue", "ld_window_kb", "ld_step",
                   "ld_r2", "pihat_min", "king_cutoff", "n_pcs",
                   "n_covariate_pcs", "chr_args", "pca_reference_populations",
                   "test_populations", "ancestry_prob_threshold",
                   "subset_population")
    thresholds <- ct |>
      filter(Parameter %in% qc_params) |>
      arrange(match(Parameter, qc_params))
    other <- ct |> filter(!Parameter %in% qc_params)
    show_table(thresholds, caption = paste0("QC parameters used for this run (",
                                            basename(cfg_path), ")."))
    collapsible("All inputs from the config file",
                sprintf("%-45s %s", other$Parameter, other$Value))
  } else {
    absent("The `jsonlite` package is not installed, so the config file could not be parsed.")
  }
} else {
  note(paste("No config file was passed. Re-render with",
             "`-P config_json=config/genotype_qc_preimputation_inputs.json`",
             "to document the thresholds this run used."))
}

cat("\n**Report generated:** ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    " · **R:** ", R.version.string,
    " · **Quarto params:** `results_dir=", params$results_dir, "`",
    if (nzchar(params$prefix)) paste0(", `prefix=", params$prefix, "`") else "",
    "\n\n", sep = "")
```

```{r sessioninfo}
#| results: asis

collapsible("R session info",
            capture.output(sessionInfo()))
```
